---
title: "Design paitent-specific logic-gated CARs for precision oncology (for in-house cohort 1)"
author: "Tiangen Chang tiangen.chang@nih.gov"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---




# load required package
```{r, include=FALSE}

library(Seurat)
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
library(AnnotationDbi) 
library(org.Hs.eg.db)
library(Matrix)
library(pheatmap)
library(RColorBrewer)
library(viridis)
library(grid)
library(gridExtra)
library(tibble)

```

# set input and output directories & parameters
```{r}

data_dir <- paste0('/02.Input/')
processed_dir <- paste0('/03.Results/')
fig_dir = paste0('/03.Results/Figures/')
surface_genes = read.csv(file = "/02.Input/Constants/cell_surface_genes_wgene_and_clinical_uniqueStandardNames.txt", header = T) 
surface_genes = as.character(surface_genes[,1])

```

# colors and constants
```{r}

# Define custom colors for each group
group_colors <- c(
  "Personalized LogiCAR" = "#08306b",  # Deep blue
  "General LogiCAR" = "#6baed6",  # light blue
  "Previous computational" = "#006d2c",      # Deep green
  "Clinical" = "#cb181d",  # Deep red
)

```

# Get the max personalized efficacy for each patient in Stefan cohort 
```{r}

## patient IDs
result_df2 = read.csv(file = paste0(data_dir,"17BRCAdatasets_AllPatientsLevel_CAR_efficacy_v2.csv"), check.names = F, row.names = 1)
stefan_Cols <- grepl("Stefan", result_df2$dataset_patient)
Stefan_patients <- result_df2$patientID[stefan_Cols]

Stefan_patients <- c("AA_10", "AA_11", "AA_13", "AA_14", "AA_15", "AA_16", "AA_18", "AA_19", "AA_2", "AA_21", "AA_22", "AA_23", "AA_24", "AA_25", "AA_27", "AA_28", "AA_3", "AA_30", "AA_31", "AA_32", "AA_4", "AA_5", "AA_6", "AA_7", "AA_8", "AA_9", "Af_1", "Af_10", "Af_11", "Af_12", "Af_13", "Af_15", "Af_16", "Af_17", "Af_18", "Af_19", "Af_2", "Af_21", "Af_23", "Af_24", "Af_25", "Af_3", "Af_5", "Af_6", "Af_7", "Af_8", "Af_9", "EA_1", "EA_10", "EA_11", "EA_12", "EA_13", "EA_14", "EA_15", "EA_16", "EA_18", "EA_19", "EA_20", "EA_21", "EA_24", "EA_25", "EA_3", "EA_4", "EA_5", "EA_6", "EA_9")

# Initialize an empty data frame to store results
result_df <- data.frame(
  patientID = character(),
  max_score = numeric(),
  stringsAsFactors = FALSE
)

# Loop through each patient ID
for (pp in Stefan_patients) {
  # Construct the result directory path
  result_dir_path_temp <- paste0(
    "/01.Scripts/Triplet_CART_designer/output/", 
    "Stefan_", pp, "_allTumorCells/"
  )
  # Read the doublets file
  doublets_file <- paste0(result_dir_path_temp, "Stefan_", pp, "_allTumorCells_optimal_doublets_merged.txt")
  if (file.exists(doublets_file)) {
    max_score_doublets <- read.csv(doublets_file, header = TRUE, row.names = NULL, sep = "\t")
    max_score_doublets <- max_score_doublets[1, 3]  # Extract the value from the first row, third column
  } else {
    warning(paste("Doublets file not found for patient:", pp))
    max_score_doublets <- NA
  }
  # Read the triplets file
  triplets_file <- paste0(result_dir_path_temp, "Stefan_", pp, "_allTumorCells_optimal_triplets_merged.txt")
  if (file.exists(triplets_file)) {
    max_score_triplets <- read.csv(triplets_file, header = TRUE, row.names = NULL, sep = "\t")
    max_score_triplets <- max_score_triplets[1, 3]  # Extract the value from the first row, third column
  } else {
    warning(paste("Triplets file not found for patient:", pp))
    max_score_triplets <- NA
  }
  # Calculate the maximum score
  max_score <- max(max_score_doublets, max_score_triplets, na.rm = TRUE)
  # Append the result to the data frame
  result_df <- rbind(result_df, data.frame(patientID = pp, max_score = max_score, stringsAsFactors = FALSE))
}

write.csv(result_df, file = "/03.Results/Stefan_cohort_personalized_efficacy_66patients.csv", quote = FALSE, row.names = F)

```

# TableS3. top 50 personalized efficacy for each patient in Stefan cohort 
```{r}

Stefan_patients <- c("AA_10", "AA_11", "AA_13", "AA_14", "AA_15", "AA_16", "AA_18", "AA_19", "AA_2", "AA_21", "AA_22", "AA_23", "AA_24", "AA_25", "AA_27", "AA_28", "AA_3", "AA_30", "AA_31", "AA_32", "AA_4", "AA_5", "AA_6", "AA_7", "AA_8", "AA_9", "Af_1", "Af_10", "Af_11", "Af_12", "Af_13", "Af_15", "Af_16", "Af_17", "Af_18", "Af_19", "Af_2", "Af_21", "Af_23", "Af_24", "Af_25", "Af_3", "Af_5", "Af_6", "Af_7", "Af_8", "Af_9", "EA_1", "EA_10", "EA_11", "EA_12", "EA_13", "EA_14", "EA_15", "EA_16", "EA_18", "EA_19", "EA_20", "EA_21", "EA_24", "EA_25", "EA_3", "EA_4", "EA_5", "EA_6", "EA_9")

# Initialize an empty data frame to store results
result_df <- data.frame(
  patientID = character(),
  Gates = character(),
  Genes = character(),
  Score = numeric(),
  stringsAsFactors = FALSE
)

# Loop through each patient ID
for (pp in Stefan_patients) {
  # Construct the result directory path
  result_dir_path_temp <- paste0(
    "/01.Scripts/Triplet_CART_designer/output/", 
    "Stefan_", pp, "_allTumorCells/"
  )
  # Read the doublets file
  doublets_file <- paste0(result_dir_path_temp, "Stefan_", pp, "_allTumorCells_optimal_doublets_merged.txt")
  if (file.exists(doublets_file)) {
    doublets_data <- read.csv(doublets_file, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
  } else {
    warning(paste("Doublets file not found for patient:", pp))
    doublets_data <- data.frame(Gates = character(), Genes = character(), Score = numeric(), stringsAsFactors = FALSE)
  }
  # Read the triplets file
  triplets_file <- paste0(result_dir_path_temp, "Stefan_", pp, "_allTumorCells_optimal_triplets_merged.txt")
  if (file.exists(triplets_file)) {
    triplets_data <- read.csv(triplets_file, header = TRUE, sep = "\t", stringsAsFactors = FALSE)
  } else {
    warning(paste("Triplets file not found for patient:", pp))
    triplets_data <- data.frame(Gates = character(), Genes = character(), Score = numeric(), stringsAsFactors = FALSE)
  }
  # Combine doublets and triplets data
  combined_data <- rbind(doublets_data, triplets_data)
  # Sort by Score in descending order
  combined_data <- combined_data[order(-combined_data$Score), ]
  # Select the top 50 solutions
  top_50 <- head(combined_data, 50)
  # Add patient ID to the top 50 solutions
  top_50$patientID <- pp
  # Append the result to the data frame
  result_df <- rbind(result_df, top_50)
}

write.csv(result_df, file = "/03.Results/TableS3.csv", quote = FALSE, row.names = F)

```

# calculate max efficacy of CARs for each of Stefan cohort patient using different groups of CARs
```{r}

#### load personalized best scores for Stefan cohort 66 patients
max_scores_df <- read.csv("/03.Results/Stefan_cohort_personalized_efficacy_66patients.csv")

#### load common_CARs
common_CARs_df = read.csv(file = paste0(data_dir,"17BRCAdatasets_AllPatientsLevel_CAR_efficacy_v2.csv"), check.names = F, row.names = 1)
## keep only Stefan cohort
stefan_Cols <- grepl("Stefan", common_CARs_df$dataset_patient)
common_CARs_df <- common_CARs_df[stefan_Cols, ]
common_CARs_df <- merge(common_CARs_df, max_scores_df, by = "patientID")
patient_subtype = paste0(common_CARs_df$patientID, "-", common_CARs_df$BRCAsubtype)
common_CARs_df$dataset = NULL
common_CARs_df$dataset_patient = NULL
common_CARs_df$patientID = NULL
common_CARs_df$BRCAsubtype = NULL

common_CARs_df <- data.frame(t(common_CARs_df))
colnames(common_CARs_df) <- patient_subtype
common_CARs_df$LogiCARs_display <- rownames(common_CARs_df)

### map CARs to different groups
candidate_CARs_df = read.csv(paste0("/01.Scripts/Triplet_CART_designer/output/safety90/filtered_top_logiCARs_efficacy_safety_17Datasets.csv"))
candidate_CARs_Groups_df <- candidate_CARs_df[c("LogiCARs_display", "Group2")]
grouped_CARs_df <- merge(common_CARs_df, candidate_CARs_Groups_df, by = "LogiCARs_display", all.x = T)
grouped_CARs_df$Group2[is.na(grouped_CARs_df$Group2)] <- "Personalized LogiCAR"
grouped_CARs_df$mean_efficacy <- rowMeans(grouped_CARs_df[2:(sum(stefan_Cols)+1)])

### remove targets that overlap between logiCAR designer and previous methods
grouped_CARs_df <- grouped_CARs_df[!grepl("\\*", grouped_CARs_df$LogiCARs_display), ]
grouped_CARs_df$Group2[grouped_CARs_df$Group2 %in% c("LogiCAR (N = 3)", "LogiCAR (N = 2)", "LogiCAR (N = 1)")] = "Best general LogiCAR"
grouped_CARs_df$Group2[grouped_CARs_df$Group2 %in% c("Kwon et al.")] = "Best Kwon et al."
grouped_CARs_df$Group2[grouped_CARs_df$Group2 %in% c("Dannenfelser et al.")] = "Best Dannenfelser et al."
grouped_CARs_df$Group2[grouped_CARs_df$Group2 %in% c("Clinical (BRCA)")] = "Best clinical (BRCA)"
grouped_CARs_df$Group2[grouped_CARs_df$Group2 %in% c("Clinical (other tumors)")] = "Best clinical (other tumors)"

# Get rows with max mean_efficacy for each Group
max_efficacy_rows <- grouped_CARs_df %>%
  group_by(Group2) %>%
  filter(mean_efficacy == max(mean_efficacy, na.rm = TRUE)) %>%
  ungroup()

max_score_vec = as.numeric(max_efficacy_rows[max_efficacy_rows$LogiCARs_display=="max_score", 2:67])
mean(max_score_vec > 0.99)
mean(max_score_vec > 0.66)
mean(max_score_vec > 0.27)

```


# Figure 5D. barplot showing enhancement of efficacy using personalized CARs
```{r}

# Extract patient names and subtypes from column names
patient_data <- max_efficacy_rows %>%
  pivot_longer(cols = 2:67, names_to = "patient_subtype", values_to = "efficacy") %>%
  separate(patient_subtype, into = c("patient", "subtype"), sep = "-") %>%
  mutate(subtype = ifelse(subtype == "LUMINAL A", "Luminal A",
                          ifelse(subtype == "LUMINAL B", "Luminal B", subtype)))

patient_data$subtype[patient_data$subtype == "HER2"] = "HER2+"
patient_data$subtype = factor(patient_data$subtype, levels = c("Luminal A", "Luminal B", "HER2+", "TNBC"))

# Filter data for bar heights (Personalized LogiCAR)
bar_data <- patient_data %>%
  filter(Group2 == "Personalized LogiCAR")

# Filter data for scatter points (other groups)
scatter_data <- patient_data %>%
  filter(Group2 != "Personalized LogiCAR")

# Order patients within each subtype by efficacy
bar_data <- bar_data %>%
  group_by(subtype) %>%
  arrange(desc(efficacy), .by_group = TRUE) %>%
  ungroup() %>%
  mutate(patient = factor(patient, levels = unique(patient)))

# Define colors for subtypes
subtype_palette <- colorRampPalette(brewer.pal(8, "Set2"))(6)
unique_subtypes <- c("Luminal A", "Luminal B", "HER2+", "HR+", "TNBC", "Metastatic")
subtype_colors <- setNames(subtype_palette, unique_subtypes)

# Define colors and shapes for scatter groups
scatter_colors <- c(
  "Best general LogiCAR" = "#2171b5",  # Dark blue
  "Best Kwon et al." = "#006d2c",      # Deep green
  "Best Dannenfelser et al." = "#41ab5d",  # Light green
  "Best clinical (BRCA)" = "#cb181d",  # Deep red
  "Best clinical (other tumors)" = "#fb6a4a"  # Light red (pink)
)

# Define shapes for scatter groups
scatter_shapes <- c(
  "Best general LogiCAR" = 16,  # Circle
  "Best Kwon et al." = 17,      # Triangle
  "Best Dannenfelser et al." = 15,  # Square
  "Best clinical (BRCA)" = 18,  # Diamond
  "Best clinical (other tumors)" = 4  # Cross (X)
)

# Ensure Group2 is a factor with the correct order
scatter_data$Group2 <- factor(scatter_data$Group2, levels = c("Best clinical (other tumors)", "Best clinical (BRCA)", "Best Dannenfelser et al.", "Best Kwon et al.", "Best general LogiCAR"))

# Save the plot as a PDF
pdf(paste0(fig_dir, "Figure5D_PatientLevel_MaxEfficacy_barplot_Stefan.pdf"), width = 10, height = 4.2)
# Create the plot
ggplot() +
  # Add bars
  geom_bar(
    data = bar_data, 
    aes(x = patient, y = efficacy * 100, fill = subtype), 
    stat = "identity", position = position_dodge(width = 0.7), width = 0.7
  ) +
  # Add scatter points (aligned to the middle of each bar)
  geom_point(
    data = scatter_data, 
    aes(x = patient, y = efficacy * 100, color = Group2, shape = Group2), 
    size = 3, 
    position = position_nudge(x = 0)  # Ensure points stay centered
  ) +
  # Add dashed grey horizontal lines
  geom_hline(yintercept = c(27, 66, 99), linetype = "dashed", color = "grey50", linewidth = 0.5) +
  # Customize colors and shapes
  scale_fill_manual(values = subtype_colors, name = "Subtype") +
  scale_color_manual(values = scatter_colors, name = "Group") +
  scale_shape_manual(values = scatter_shapes, name = "Group") +
  # Fix y-axis range
  scale_y_continuous(limits = c(0, 100)) +
  # Theme adjustments
  theme(
    panel.grid = element_blank(), 
    panel.background = element_rect(fill = "white"),
    axis.line.y = element_line(color = "black"),
    axis.line.x = element_line(color = "black"),
    axis.ticks.y = element_line(color = "black"),
    axis.ticks.x = element_blank(),  # Hide x-axis ticks
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, color = "black"),  # Rotate x-axis labels
    axis.title.x = element_blank(),  # Hide x-axis title
    axis.text.y = element_text(color = "black"),
    panel.border = element_blank(), 
    legend.position = "right"  # Legend position
  ) +
  labs(x = "Patient", y = "Efficacy (%)", title = "")
dev.off()

```

# Figure 5E. average efficacy in different subtypes
```{r}

efficay_df = as.data.frame(max_efficacy_rows[,2:67])
rownames(efficay_df) = max_efficacy_rows$Group2
efficay_df <- efficay_df[c("Best clinical (other tumors)", "Best clinical (BRCA)", "Best Dannenfelser et al.", "Best Kwon et al.", "Best general LogiCAR", "Personalized LogiCAR"),]
efficay_df[1:6,1:6]

row_medians <- rowMedians(as.matrix(efficay_df[, sapply(efficay_df, is.numeric)]))
row_means <- rowMeans(as.matrix(efficay_df[, sapply(efficay_df, is.numeric)]))

# Reshape the data into long format
efficay_long <- efficay_df %>%
  rownames_to_column(var = "Group") %>%  # Convert row names to a column
  pivot_longer(cols = -Group, names_to = "Patient", values_to = "Efficacy") %>%
  mutate(Group = factor(Group, levels = c("Best clinical (other tumors)", "Best clinical (BRCA)", 
                                          "Best Dannenfelser et al.", "Best Kwon et al.", 
                                          "Best general LogiCAR", "Personalized LogiCAR")))

# Define colors for groups
group_colors <- c(
  "Best clinical (other tumors)" = "#fb6a4a",  # Light red (pink)
  "Best clinical (BRCA)" = "#cb181d",          # Deep red
  "Best Dannenfelser et al." = "#41ab5d",      # Light green
  "Best Kwon et al." = "#006d2c",              # Deep green
  "Best general LogiCAR" = "#2171b5",          # Dark blue
  "Personalized LogiCAR" = "#08306b"             # Black
)

# Create the boxplot with scatter points
pdf(paste0(fig_dir, "Figure5E_Patient_Efficacy_Stefan_Personalized_vs_general_CARs_boxplot.pdf"), width = 3*0.8, height = 4*0.8)
ggplot(efficay_long, aes(x = Group, y = Efficacy * 100, fill = Group)) +
  # Add scatter points
  geom_jitter(aes(color = Group), width = 0.2, alpha = 0.7, size = 2) +
  # Add boxplots
  geom_boxplot(outlier.shape = NA, color = "black", alpha = 0.7) +
  # Customize colors
  scale_fill_manual(values = group_colors, name = "Group") +
  scale_color_manual(values = group_colors, name = "Group") +
  # Fix y-axis range
  scale_y_continuous(limits = c(0, 100)) +
  theme(
    panel.grid = element_blank(), 
    panel.background = element_rect(fill = "white"),
    axis.line.y = element_line(color = "black"),
    axis.line.x = element_line(color = "black"),
    axis.ticks.y = element_line(color = "black"),
    axis.ticks.x = element_line(color = "black"),
    panel.border = element_blank(), 
    axis.text.y = element_text(color = "black"),  # Assign colors to y-axis tick labels
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5, color = "black"),  
    legend.position = "none"  # Legend position
  ) + 
  labs(
    x = "",
    y = "Efficacy (%)",
    title = ""
  )
dev.off()

a = as.numeric(efficay_df["Personalized LogiCAR", ])
b = as.numeric(efficay_df["Best Dannenfelser et al.", ]) # "Best clinical (other tumors)", "Best clinical (BRCA)", "Best Dannenfelser et al.", "Best Kwon et al.", "Best general LogiCAR"
wilcox.test(a,b)
wilcox.test(a,b, paired = T)

```
