---
title: "Evaluate the efficacy of potential CARs and logic-gated CARs across Discovery and External-Validation datasets"
author: "Tiangen Chang tiangen.chang@nih.gov"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



# Connect to server
```{bash}
ssh changt7@biowulf.nih.gov
sinteractive --mem=512g --cpus-per-task=64 --time=24:00:00
# --gres=lscratch:1024
module load R/
R

```


# load required package
```{r, include=FALSE}

#devtools::install_github("mojaveazure/seurat-disk")
#devtools::install_github('satijalab/seurat-data')
library(SeuratData)
library(SeuratDisk)

library(caret)
library(Seurat)
library(dplyr)
library(ggplot2)
library(dittoSeq)
library(SeuratWrappers)
library(patchwork)
library(RColorBrewer)
library(stringr)
library(tidyr)
library(data.table)
library(Matrix)

#install.packages("showtext")
library(showtext) # this package enables R to show special characters like ≤
# Add a font and enable `showtext`
font_add(family = "Arial", regular = "path/to/arial.ttf")  # Replace with the path to your Arial font file
showtext_auto()

```


####################
# check all below code, ensure '-' in gene name has been converted to '_' during processing
# '-' is used only in display!!!
####################



# set input and output directories & parameters
```{r}

# server
data_dir <- paste0('/data/Lab_ruppin/tiangen/CancerProject/08.Sanna_CART/02.Input/')
processed_dir <- paste0('/data/Lab_ruppin/tiangen/CancerProject/08.Sanna_CART/03.Results/')
fig_dir = paste0('/data/Lab_ruppin/tiangen/CancerProject/08.Sanna_CART/03.Results/Figures/')
#surface_genes = read.csv(file = "/data/Lab_ruppin/tiangen/CancerProject/08.Sanna_CART/02.Input/Constants/intersection_and_clinical.txt", header = F)
#surface_genes = read.csv(file = "/data/Lab_ruppin/tiangen/CancerProject/08.Sanna_CART/02.Input/Constants/cell_surface_genes_wgene_and_clinical.txt", header = T) # 2786 genes
surface_genes = read.csv(file = "/data/Lab_ruppin/tiangen/CancerProject/08.Sanna_CART/02.Input/Constants/cell_surface_genes_wgene_and_clinical_uniqueStandardNames.txt", header = T) # 2758 genes
surface_genes = as.character(surface_genes[,1])

# local
data_dir <- paste0('/Users/changt7/Documents/00.PostDocWork/CancerResearch/01.Projects/00.Collaborating_projects/08.Sanna_CART/02.Data/')
fig_dir = paste0('/Users/changt7/Documents/00.PostDocWork/CancerResearch/01.Projects/00.Collaborating_projects/08.Sanna_CART/03.Results/Figures/')



my.colors <- c(
  "#E41A1C", # Bright Red
  "#377EB8", # Bright Blue
  "#4DAF4A", # Green
  "#984EA3", # Purple
  "#FF7F00", # Orange
  "#FFFF33", # Yellow
  "#A65628", # Brown
  "#F781BF", # Pink
  "#999999", # Gray
  "#66C2A5", # Turquoise
  "#FC8D62", # Salmon
  "#8DA0CB", # Light Blue
  "#E78AC3", # Light Pink
  "#A6D854", # Lime Green
  "#FFD92F", # Light Yellow
  "#E5C494", # Beige
  "#7570B3", # Blue-Violet
  "#5B1A18", # Deep Brown/Maroon
  "#8E6E53", # Taupe
  "#CAB2D6", # Light Purple
  "#1B9E77"  # Teal
)

# Define custom colors for each group
group_colors <- c(
  "LogiCAR (N = 5)" = "#08306b",  # Deep blue
  "LogiCAR (N = 4)" = "#2171b5",  # Dark blue
  "LogiCAR (N = 3)" = "#4292c6",  # Medium blue
  "LogiCAR (N = 2)" = "#6baed6",  # Light blue
  "LogiCAR (N = 1)" = "#9ecae1",  # Very light blue
  "Kwon et al." = "#006d2c",      # Deep green
  "Dannenfelser et al." = "#41ab5d",  # Light green
  "Clinical (BRCA)" = "#cb181d",  # Deep red
  "Clinical (other tumors)" = "#fb6a4a"  # Light red (pink)
)

```

# load common functions
```{r}

source("00.Common_Functions.R")

```


# (Optional) check HPA normal cells and public tumor cells input data for LogiCAR Designer
```{r}

geneExpr_tumor_df <- fread(paste0(data_dir,"/15BRCAdatasets_tumorCells_LogiCAR_wgene_and_clinical.txt"), sep = "\t", header = FALSE)
dim(geneExpr_tumor_df)
geneExpr_tumor_df[1:5,1:5]

geneExpr_HPA_df <- fread(paste0(data_dir,"/HPA/version24_20241022/HPA_v24_1kCellsPerTissue_LogiCAR_cell_surface_genes_wgene_and_clinical.txt"), sep = "\t", header = FALSE)
dim(geneExpr_HPA_df)
geneExpr_HPA_df[1:5,1:5]

```


# (run once only!) read in and process top N-gene logic-gated CARs identified by LogiCAR Designer to standard format
```{r}

#top_logiCARs_data_dir = "/Users/changt7/Documents/00.PostDocWork/CancerResearch/01.Projects/00.Collaborating_projects/08.Sanna_CART/01.Scripts/Triplet_CART_designer/output/safety80/"
top_logiCARs_data_dir = "/Users/changt7/Documents/00.PostDocWork/CancerResearch/01.Projects/00.Collaborating_projects/08.Sanna_CART/01.Scripts/Triplet_CART_designer/output/safety90/"

top_logiCARs_files_suffix = c("singlets", "doublets_merged", "triplets_merged", "quadruples_merged", "quintuplets_merged")
keep_topN = 15

### extract top solutions with considering duplications (with N-1 genes the same)
top_logiCARs_summary_df = data.frame()
top_logiCARs_uniqueGenes = c()
for (fn in top_logiCARs_files_suffix) {
  print(paste(fn, "in processing ..."))
  top_logiCARs_df = read.csv(file = paste0(top_logiCARs_data_dir, "AllDatasets_EachPatient500Cells_optimal_",fn,".txt"), sep = "\t", header = T)
  top_logiCARs_dt = as.data.table(top_logiCARs_df)
  # Remove brackets and split genes into a vector
  top_logiCARs_dt$Genes <- str_replace_all(top_logiCARs_dt$Genes, "\\[|\\]|\\'", "")
  top_logiCARs_dt[, GenesSplit := strsplit(Genes, ", ")]
  current_list = list()
  current_list[[1]] <- list(top_logiCARs_dt$GenesSplit[[1]], round(top_logiCARs_dt$Score[[1]],2), top_logiCARs_dt$LogicGates[[1]])
  current_list_len = 1
  for (row_i in 2:nrow(top_logiCARs_dt)){
    flag = 1
    if (fn != "singlets"){
      for (item_j in 1:current_list_len){
        Score_temp = round(top_logiCARs_dt$Score[[row_i]],2)
        diff_num = length(setdiff(current_list[[item_j]][[1]], top_logiCARs_dt$GenesSplit[[row_i]]))
        if ((diff_num < 2) & (Score_temp == current_list[[item_j]][[2]])){
          flag = 0
          break
        }
      }
    }
    if (flag == 1){
      current_list[[current_list_len+1]] <- list(top_logiCARs_dt$GenesSplit[[row_i]], Score_temp, top_logiCARs_dt$LogicGates[[row_i]])
      current_list_len = current_list_len + 1
      if (current_list_len == keep_topN){
        break
      }
    }
  }
  rows_used = row_i
  print(paste(fn, rows_used))
  #print(current_list)
  # Apply the conversion function to the data to get standard representation of logiCARs
  top_logiCARs_dt[, LogiCARs := mapply(convert_logic, LogicGates, Genes)]
  
  unique_genes <- unique(unlist(top_logiCARs_dt$GenesSplit[1:rows_used]))
  unique_genes <- gsub("_not", "", unique_genes)
  result_df_temp <- top_logiCARs_dt[1:rows_used, c("Genes", "LogiCARs", "Score", "GenesSplit")]
  geneNum = length(top_logiCARs_dt$GenesSplit[[1]])
  result_df_temp$geneNum = geneNum
  top_logiCARs_summary_df = rbind(top_logiCARs_summary_df, result_df_temp)
  top_logiCARs_uniqueGenes = unique(c(top_logiCARs_uniqueGenes, unique_genes))
  #print(current_list)
}

## convert - in gene names to _
top_logiCARs_summary_df$LogiCARs = gsub("-", "_", top_logiCARs_summary_df$LogiCARs)

# Print the results
dim(top_logiCARs_summary_df)
length(top_logiCARs_uniqueGenes)

```

# load and process sampled tumor and normal cell RNA expression data (as LogiCAR Designer input)
```{r}

### load HPA RNA Discovery data  31000 cells *  5448/2 genes
data_HPA_RNA = fread(file = paste0("/Users/changt7/Documents/00.PostDocWork/CancerResearch/01.Projects/00.Collaborating_projects/08.Sanna_CART/02.Input/HPA/version24_20241022/rna/HPA_LogiCAR_cell_surface_genes_wgene_and_clinical.txt"), sep = "\t", header = F) 
data_HPA_RNA[1:5,1:5]
data_HPA_RNA_onlyData = data_HPA_RNA[3:nrow(data_HPA_RNA),,drop=F]
data_HPA_RNA_onlyData[1:5,1:5]
data_HPA_RNA_geneNAs = data_HPA_RNA_onlyData$V1
data_HPA_RNA_onlyData$V1 = NULL

# Step 1: Convert the data.table to a numeric matrix
data_HPA_RNA_onlyData <- as.matrix(data_HPA_RNA_onlyData)
rownames(data_HPA_RNA_onlyData) <- data_HPA_RNA_geneNAs
data_HPA_RNA_onlyData <- apply(data_HPA_RNA_onlyData, 2, as.numeric)  # Convert to numeric
# Step 2: Convert the numeric matrix to a sparse matrix
data_HPA_RNA_onlyData <- Matrix(data_HPA_RNA_onlyData, sparse = TRUE)
# Step 3: Transpose the sparse matrix
data_HPA_RNA_onlyData <- t(data_HPA_RNA_onlyData)
dim(data_HPA_RNA_onlyData)
colnames(data_HPA_RNA_onlyData) <- data_HPA_RNA_geneNAs
data_HPA_RNA_onlyData_negated <- data_HPA_RNA_onlyData == 0
data_HPA_RNA_onlyData_negated <- as(data_HPA_RNA_onlyData_negated, "CsparseMatrix")
colnames(data_HPA_RNA_onlyData_negated) = paste0(data_HPA_RNA_geneNAs, "_not")
data_HPA_RNA_final_matrix <- cbind(data_HPA_RNA_onlyData, data_HPA_RNA_onlyData_negated)
dim(data_HPA_RNA_final_matrix)
data_HPA_RNA_final_matrix[1:5,1:5]
## convert - in gene names to _
colnames(data_HPA_RNA_final_matrix) = gsub("-", "_", colnames(data_HPA_RNA_final_matrix))


#### load tumor data (public datasets Discovery) 72432 cells *  5450/2 genes
data_tumor = fread("/Users/changt7/Documents/00.PostDocWork/CancerResearch/01.Projects/00.Collaborating_projects/08.Sanna_CART/02.Input/15BRCAdatasets_tumorCells_LogiCAR_wgene_and_clinical.txt", sep = "\t", header = F)
data_tumor[1:5,1:5]
data_tumor_onlyData = data_tumor[4:nrow(data_tumor),,drop=F]
data_tumor_onlyData[1:5,1:5]
data_tumor_geneNAs <- data_tumor_onlyData$V1
data_tumor_onlyData$V1 = NULL

# Step 1: Convert the data.table to a numeric matrix
data_tumor_onlyData <- as.matrix(data_tumor_onlyData)
rownames(data_tumor_onlyData) <- data_tumor_geneNAs
data_tumor_onlyData <- apply(data_tumor_onlyData, 2, as.numeric)  # Convert to numeric
# Step 2: Convert the numeric matrix to a sparse matrix
data_tumor_onlyData <- Matrix(data_tumor_onlyData, sparse = TRUE)
# Step 3: Transpose the sparse matrix
data_tumor_onlyData <- t(data_tumor_onlyData)
dim(data_tumor_onlyData)
colnames(data_tumor_onlyData) <- data_tumor_geneNAs
data_tumor_onlyData_negated <- data_tumor_onlyData == 0
data_tumor_onlyData_negated <- as(data_tumor_onlyData_negated, "CsparseMatrix")
colnames(data_tumor_onlyData_negated) = paste0(data_tumor_geneNAs, "_not")
data_tumor_final_matrix <- cbind(data_tumor_onlyData, data_tumor_onlyData_negated)
dim(data_tumor_final_matrix)
data_tumor_final_matrix[1:5,1:5]
## convert - in gene names to _
colnames(data_tumor_final_matrix) = gsub("-", "_", colnames(data_tumor_final_matrix))


```

# (run for once only!!) filtering top N-gene logic-gated CARs identified by LogiCAR Designer by safety
# filter identified top logiCARs by safety (for same efficacy and N-1 same gene logiCARs, only keep one highest safety)
```{r}

# efficacy RNA
efficacy_res_list <- calcEquivalentExpression2(top_logiCARs_summary_df$LogiCARs, data_tumor_final_matrix) 
length(efficacy_res_list)
length(efficacy_res_list[[1]])

# safety RNA
safety_res_list <- calcEquivalentExpression2(top_logiCARs_summary_df$LogiCARs, data_HPA_RNA_final_matrix) 
length(safety_res_list)
length(safety_res_list[[1]])

result_df = data.frame(LogiCARs = as.character(),
                       Efficay = as.numeric(),
                       Safety = as.numeric()
                       )
# Loop through the names in safety_res_list
for (le in names(safety_res_list)) {
  # Calculate Efficacy and Safety
  Efficacy <- mean(efficacy_res_list[[le]])
  Safety <- 1 - mean(safety_res_list[[le]])
  # Append the results as a new row to the data frame
  result_df <- rbind(result_df, data.frame(LogiCARs = le, Efficacy = Efficacy, Safety = Safety))
}


top_logiCARs_summary_df_scored = merge(top_logiCARs_summary_df, result_df, by = "LogiCARs")

### confirm that Score = Efficacy
plot(top_logiCARs_summary_df_scored$Score, top_logiCARs_summary_df_scored$Efficacy)


############################### filter LogiCARs to keep top N non-redundant targets
has_n_minus_1_common_genes <- function(genes1, genes2) {
  diff_genes <- setdiff(genes1, genes2)
  return((length(diff_genes) <2) & length(genes1) > 1)
}


# ## Step1 Filter rows: Remove duplicates based on Score, keeping the highest Safety
# filtered_top_logiCARs <- top_logiCARs_summary_df_scored[
#   order(-Score, -Safety),  # Sort by Score (descending) and Safety (descending)
#   .SD[!duplicated(Score)], # Remove duplicates based on Score, keeping the highest Safety
#   by = geneNum             # Group by geneNum
# ]

# Reduce the resolution of Score
top_logiCARs_summary_df_scored[, Score := round(Score, 2)]

## Step 2: Deduplicate rows by grouping rows with N-1 genes the same and the same Score
filtered_rows <- list()
for (gene_num in unique(top_logiCARs_summary_df_scored$geneNum)) {
  group <- top_logiCARs_summary_df_scored[geneNum == gene_num]
  keep_rows <- !logical(nrow(group)) # all TRUE
  for (i in 1:(nrow(group)-1)) {
    if (keep_rows[i]) {
      for (j in (i + 1):nrow(group)) {
        if (keep_rows[j] &&
            group$Score[i] == group$Score[j] &&
            has_n_minus_1_common_genes(group$GenesSplit[[i]], group$GenesSplit[[j]])) {
            if (group$Safety[i] >= group$Safety[j]) {
              keep_rows[j] <- FALSE
            } else {
              keep_rows[i] <- FALSE
              break
            }
          }
        }
      }
    }
  print(paste(gene_num, sum(keep_rows)))
  filtered_rows[[as.character(gene_num)]] <- group[keep_rows]
}

filtered_top_logiCARs <- rbindlist(filtered_rows)

## Step 3: Keep only the top 10 rows per geneNum after deduplication
filtered_top_logiCARs <- filtered_top_logiCARs[
  order(-Score, -Safety),
  head(.SD, 10),
  by = geneNum
]


filtered_top_logiCARs = filtered_top_logiCARs[order(-filtered_top_logiCARs$Efficacy), ]

filtered_top_logiCARs <- filtered_top_logiCARs[
  order(filtered_top_logiCARs$geneNum, -filtered_top_logiCARs$Efficacy),
]

save(filtered_top_logiCARs, file = "/Users/changt7/Documents/00.PostDocWork/CancerResearch/01.Projects/00.Collaborating_projects/08.Sanna_CART/01.Scripts/Triplet_CART_designer/output/safety90/filtered_top_logiCARs.RData")
#save(filtered_top_logiCARs, file = "/Users/changt7/Documents/00.PostDocWork/CancerResearch/01.Projects/00.Collaborating_projects/08.Sanna_CART/01.Scripts/Triplet_CART_designer/output/safety80/filtered_top_logiCARs.RData")

```


# load and process full tumor and normal cell expression data (all cells, filtered with surface genes only)
```{r}


### load HPA RNA data
# server
#seu_HPA_RNA = readRDS(file = paste0("/data/Lab_ruppin/tiangen/CancerProject/08.Sanna_CART/02.Input/HPA/version24_20241022/seu_HPA_v24.rds"))  # 20025 features across 689601 samples within 1 assay
# local
seu_HPA_RNA = readRDS(file = paste0("/Users/changt7/Documents/00.PostDocWork/CancerResearch/01.Projects/00.Collaborating_projects/08.Sanna_CART/02.Input/HPA/version24_20241022/rna/seu_HPA_v24.rds"))  # 20025 features across 689601 samples within 1 assay
genes_HPA_RNA = rownames(seu_HPA_RNA)
seu_HPA_RNA_counts_matrix <- GetAssayData(seu_HPA_RNA, slot = "counts")


#### load tumor data (public datasets Discovery)
## server
#seu_tumor_15public = readRDS("/data/Lab_ruppin/tiangen/CancerProject/08.Sanna_CART/02.Input/seu_15BRCAdatasets_tumorCells_surfaceGenes_4MetaCol.rds")
## local
seu_tumor_15public = readRDS("/Users/changt7/Documents/00.PostDocWork/CancerResearch/01.Projects/00.Collaborating_projects/08.Sanna_CART/02.Input/seu_15BRCAdatasets_AllTumorCells_ProteinCodingGenes_4MetaCol.rds")
colnames(seu_tumor_15public@meta.data)
unique_15public_datasets = unique(seu_tumor_15public$dataset)
aligned_table <- table(seu_tumor_15public$dataset)[match(unique_15public_datasets, names(table(seu_tumor_15public$dataset)))]
tumor_15public_break_points <- c(0, cumsum(as.numeric(aligned_table)))
genes_15public = rownames(seu_tumor_15public)
tumor_15public_mat <- GetAssayData(seu_tumor_15public, layer = "RNA", slot = "counts")


#### load tumor data (in-house datasets External)
data_dir_Stefan = "/Users/changt7/Documents/00.PostDocWork/CancerResearch/01.Projects/00.Collaborating_projects/08.Sanna_CART/02.Input/Stefan_snRNAseq/"
data_dir_GSE246613 = "/Users/changt7/Documents/00.PostDocWork/CancerResearch/01.Projects/00.Collaborating_projects/08.Sanna_CART/02.Input/GSE246613/"
# seu_Stefan_tumor = readRDS(paste0(data_dir_Stefan, "Multiomics82_Tumor.rds"))
# seu_GSE246613 = readRDS(paste0(data_dir_GSE246613, "GSE246613_seu_Tissue.rds"))
geneExpr_mat_Stefan = readRDS(paste0(data_dir_Stefan,"Stefan_snRNAseq_proteinCodingGenes_TumorCells_sparse_matrix.rds"))
genes_Stefan = rownames(geneExpr_mat_Stefan)
geneExpr_mat_GSE246613 = readRDS(paste0(data_dir_GSE246613,"GSE246613_proteinCodingGenes_TumorCells_sparse_matrix.rds"))
genes_GSE246613 = rownames(geneExpr_mat_GSE246613)
geneExpr_mat_Stefan[1:5,1:5]
geneExpr_mat_GSE246613[1:5,1:5]

allGenes_tumor = intersect(genes_Stefan, genes_GSE246613) # 19088
allGenes_measured = intersect(intersect(genes_HPA_RNA, genes_15public), allGenes_tumor) # 18983
c("ELAPOR1", "ERBB4", "PRLR") %in% allGenes_measured


```


# extract all candidate CAR targets and genes for comparison
```{r}


clinical_targets = c("MUC1", "CD274", "PROM1", "CD70", "ROR1", "CD80", "CD86", "TNFRSF10B", "FOLR1", "MPP2", "MUC16", "PMEL", "ROR2", "KDR", "EPHA2", "PSCA", "MET", "IL13RA2", "EPCAM", "EGFR", "FOLH1", "GPC3", "ERBB2", "MSLN", "CD276", "TM4SF1", "BSG", "CEACAM5", "ALPP", "GUCY2C", "TACSTD2")
### (viable) previous computational combinations (exclude combinations with non-surface genes, i.e., genes not in our surface gene list)
### old
# Comptutional_CAR_bulk = c("FAP & EPHA2_not", "FAP & MET_not", "OR2B6 & MET_not", "SLC1A4 & SSTR1_not", "ADAM12 & KDR_not", "EPHB3 & FZD5_not",  "ADAM12 & GPC3_not", "MUC1 & LYVE1_not", "SLC4A5 & PTH1R_not", "FLVCR1 & TRHDE_not", "EPHB3 & SLC16A7_not", "EPCAM & SSTR1_not", "MUC1 & SLC4A4_not", "TNFRSF18 & SLC16A7_not", "LDLRAD3 & SSTR1_not")
#### top 10
Comptutional_CAR_bulk <- c("FAP & EPHA2_not", "FAP & MET_not", "OR2B6 & EPHA2_not", "OR2B6 & MET_not", "SLC1A4 & SSTR1_not", "ADAM12 & KDR_not", "EPHB3 & FZD5_not", "ADAM12 & GPC3_not", "MUC1 & LYVE1_not", "SLC4A5 & PTH1R_not")
#### top 20
#Comptutional_CAR_bulk <- c("FAP & EPHA2_not", "FAP & MET_not", "OR2B6 & EPHA2_not", "OR2B6 & MET_not", "SLC1A4 & SSTR1_not", "ADAM12 & KDR_not", "EPHB3 & FZD5_not", "ADAM12 & GPC3_not", "MUC1 & LYVE1_not", "SLC4A5 & PTH1R_not", "FLVCR1 & TRHDE_not", "EPHB3 & SLC16A7_not", "EPCAM & SSTR1_not", "MUC1 & SLC4A4_not", "TNFRSF18 & SLC16A7_not", "LDLRAD3 & SSTR1_not", "KREMEN2 & KL_not", "KREMEN2 & EPHA2_not", "KREMEN2 & MET_not", "KISS1R & TNFRSF9_not")

### top 5 for AND, OR, NOT gates respectively
Comptutional_CAR_sc = c("SLC39A6 & SPINT2", "TACSTD2 & TSPAN13", "CD24 & TM4SF1", "CD9 & SLC39A6", "BCAM & SPINT2",
                        "BAMBI | PRLR", "NECTIN4 | PRLR", "BAMBI | NECTIN4", "LY6K | PRLR", "PRLR | VTCN1",
                        "CLDN3 & TSPAN8_not", "SERINC2 & TSPAN8_not", "CLDN3 & PIGR_not", "PLPP2 & MUC4_not", "SERINC2 & PIGR_not"
                        )
#other = c("MUC1 | ERBB2", "ERBB4 | PRLR | ELAPOR1", "CLDN3 | ERBB4 | PRLR", "(HLA_E_not | CLDN3) & EMP3_not") # (MUC1 & ERBB2) https://link.springer.com/article/10.1007/s10875-012-9689-9
other_targets = c()

# Logic expressions
### newly identified filtered top targets by LogiCAR Designer
#load(file = "/Users/changt7/Documents/00.PostDocWork/CancerResearch/01.Projects/00.Collaborating_projects/08.Sanna_CART/01.Scripts/Triplet_CART_designer/output/safety80/filtered_top_logiCARs.RData") # variable name filtered_top_logiCARs
logiCAR_combos = load(file = "/Users/changt7/Documents/00.PostDocWork/CancerResearch/01.Projects/00.Collaborating_projects/08.Sanna_CART/01.Scripts/Triplet_CART_designer/output/safety90/filtered_top_logiCARs.RData")
logiCAR_combos <- filtered_top_logiCARs$LogiCARs


targets_vec = c(logiCAR_combos, clinical_targets, Comptutional_CAR_bulk, Comptutional_CAR_sc, other_targets)
logiCAR_combos_df = as.data.frame(filtered_top_logiCARs)[c("LogiCARs", "geneNum")]
logiCAR_combos_df$Group = "LogiCAR Designer"
Clinical_CARs_df = data.frame(LogiCARs = clinical_targets)
Clinical_CARs_df$geneNum = 1
Clinical_CARs_df$Group = "Clinical Targets"
Comptutional_CARs_df = data.frame(LogiCARs = c(Comptutional_CAR_bulk, Comptutional_CAR_sc))
Comptutional_CARs_df$geneNum = 2
Comptutional_CARs_df$Group = c(rep("Previous Comptutional (bulk)", length(Comptutional_CAR_bulk)), rep("Previous Comptutional (sc)", length(Comptutional_CAR_sc)))
Other_CARs_df <- if (length(other_targets) > 0) {
  data.frame(
    LogiCARs = other_targets,
    geneNum = NA,
    Group = "Other Targets"
  )
} else {
  data.frame(
    LogiCARs = character(0),
    geneNum = numeric(0),
    Group = character(0)
  )
}
targets_df = rbind(logiCAR_combos_df, Clinical_CARs_df, Comptutional_CARs_df, Other_CARs_df)


## get unique single genes from above vector
cleaned_combos <- gsub("[|&()]", "", targets_vec)  # Remove |, &, (, )
cleaned_combos <- gsub("_not", "", cleaned_combos)    # Remove _not
all_genes <- unlist(strsplit(cleaned_combos, "\\s+"))
unique_genes <- unique(all_genes)
unique_CARtargets_measured = unique_genes # intersect(unique_genes, allGenes_measured) # allGenes_tumor allGenes_measured
## get targets that all genes present in HPA protein measurement
# Function to check if all genes in a logic combination are in unique_CARtargets_measured
is_valid_combo <- function(combo, valid_genes) {
  # Extract genes from the logic combination
  genes <- unlist(strsplit(combo, "[|&() ]+"))  # Split by logical operators and spaces
  genes <- gsub("_not", "", genes)    # Remove _not
  genes <- genes[genes != ""]  # Remove empty strings
  # Check if all genes are in valid_genes
  all(genes %in% valid_genes)
}
filtered_indices = sapply(targets_vec, is_valid_combo, valid_genes = unique_CARtargets_measured)
filtered_targets_vec <- targets_vec[filtered_indices]
filtered_targets_df <- targets_df[filtered_indices, ]


```

# Figure 2B-C. SFigure**
# plot boxplot showing the efficacy of N-gene CARs (N = 1,2,3,4,5) on Discovery data (500 cells per patient in 15 public datasets), full Stefan, and full Simon
# plot boxplot showing the safety of N-gene CARs (N = 1,2,3,4,5) on Discovery data (1000 cells per dataset in HPA 31 public datasets)
```{r}

## identify duplicated CAR targets in different groups and remove duplicates
duplicated_CARs = names(table(filtered_targets_df$LogiCARs)[table(filtered_targets_df$LogiCARs) > 1])
filtered_targets_df[filtered_targets_df$LogiCARs %in% duplicated_CARs, ]
#filtered_targets_df <- filtered_targets_df[!duplicated(filtered_targets_df$LogiCARs), ]
#filtered_targets_df[filtered_targets_df$LogiCARs %in% duplicated_CARs, ]
filtered_targets_df$LogiCARs_display <- filtered_targets_df$LogiCARs
for (cc in duplicated_CARs) {
  # Find the indices of the current duplicated LogiCARs value
  indices <- which(filtered_targets_df$LogiCARs == cc)
  # Add "*", "**", etc., based on the occurrence count
  for (i in seq_along(indices)) {
    if (i > 1) {  # Skip the first occurrence
      filtered_targets_df$LogiCARs_display[indices[i]] <- paste0(cc, strrep("*", i - 1))
    }
  }
}


# efficacy RNA
efficacy_res_list <- calcEquivalentExpression2(filtered_targets_df$LogiCARs, data_tumor_final_matrix) 
# first para: targets_vec; unique_genes_in_HPAprotein; filtered_targets_vec; unique_clinical_targets_measured; filtered_targets_df$LogiCARs
# second para: tumor_15public_final_matrix; geneExpr_mat_GSE246613_CARs_final_matrix; geneExpr_mat_Stefan_CARs_final_matrix; data_tumor_final_matrix
length(efficacy_res_list)
length(efficacy_res_list[[1]])
names(efficacy_res_list) <- filtered_targets_df$LogiCARs_display

# safety RNA
safety_res_list <- calcEquivalentExpression2(filtered_targets_df$LogiCARs, data_HPA_RNA_final_matrix) 
length(safety_res_list)
length(safety_res_list[[1]])
names(safety_res_list) <- filtered_targets_df$LogiCARs_display


result_df = data.frame(LogiCARs_display = as.character(),
  Efficay = as.numeric(),
  Safety = as.numeric()
  )
# Loop through the names in safety_res_list
for (le in names(safety_res_list)) {
  # Calculate Efficacy and Safety
  Efficacy <- mean(efficacy_res_list[[le]])
  Safety <- 1 - mean(safety_res_list[[le]])
  # Append the results as a new row to the data frame
  result_df <- rbind(result_df, data.frame(LogiCARs_display = le, Efficacy = Efficacy, Safety = Safety))
}

# result_df = result_df[order(-result_df$Efficacy), ]
# result_df2 = result_df[order(result_df$Safety), ]
# print(result_df, width = 200)

filtered_targets_withResult_df = merge(filtered_targets_df, result_df, by = "LogiCARs_display")
dim(filtered_targets_withResult_df)



########################### boxplot
filtered_targets_withResult_df$Group2 = filtered_targets_withResult_df$Group
filtered_targets_withResult_df$Group2[filtered_targets_withResult_df$Group2 == "Clinical Targets"] = "Clinical (other tumors)"
filtered_targets_withResult_df$Group2[filtered_targets_withResult_df$LogiCARs %in% c("MUC1", "CD274", "PROM1", "CD70", "ROR1", "PSCA", "MET", "EPCAM", "EGFR", "FOLH1", "ERBB2", "MSLN", "CD276", "TACSTD2")] = "Clinical (BRCA)"
filtered_targets_withResult_df$Group2[filtered_targets_withResult_df$Group2 == "LogiCAR Designer"] = paste0("LogiCAR (N = ", filtered_targets_withResult_df$geneNum[filtered_targets_withResult_df$Group2 == "LogiCAR Designer"], ")")
filtered_targets_withResult_df$Group2[filtered_targets_withResult_df$Group2 == "Previous Comptutional (bulk)"] = "Dannenfelser et al."
filtered_targets_withResult_df$Group2[filtered_targets_withResult_df$Group2 == "Previous Comptutional (sc)"] = "Kwon et al."
#filtered_targets_withResult_df$Group2 = factor(filtered_targets_withResult_df$Group2, levels = c("LogiCAR (N = 5)", "LogiCAR (N = 4)", "LogiCAR (N = 3)", "LogiCAR (N = 2)", "LogiCAR (N = 1)", "Kwon et al.", "Dannenfelser et al.", "Clinical (BRCA)", "Clinical (other tumors)"))
filtered_targets_withResult_df$Group2 = factor(filtered_targets_withResult_df$Group2, levels = rev(c("LogiCAR (N = 5)", "LogiCAR (N = 4)", "LogiCAR (N = 3)", "LogiCAR (N = 2)", "Kwon et al.", "Dannenfelser et al.", "LogiCAR (N = 1)", "Clinical (BRCA)", "Clinical (other tumors)")))

############ Boxplot with scatter points for efficacy comparison
pdf_file <- paste0(fig_dir, "Figure2c_efficacy_on_Discovery_safety90.pdf")
fig_width <- 2.3
fig_height <- 2.7
fontSize <- 1.2
pdf(pdf_file, onefile = FALSE, width = fig_width, height = fig_height)
ggplot(filtered_targets_withResult_df, aes(x = Group2, y = Efficacy * 100)) +
  geom_boxplot(outlier.shape = NA, color = "black", fill = NA) +  # Black boxplot with no fill
  geom_jitter(width = 0.2, size = 1, alpha = 1, aes(color = Group2)) +  # Scatter points with group colors
  scale_color_manual(values = group_colors) +  # Apply custom colors to scatter points
  labs(
    title = "",
    x = "",
    y = "Efficacy (%)"
  ) +
  scale_y_continuous(limits = c(0, 100)) +  # Set y-axis range from 0 to 100
  theme(
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "white"),
    axis.line.y = element_line(color = "black"),
    axis.line.x = element_line(color = "black"),
    axis.ticks.y = element_line(color = "black"),
    axis.ticks.x = element_line(color = "black"),
    axis.text.y = element_text(color = "black"),
    panel.border = element_blank(),
    axis.text.x = element_text(color = "black", angle = 60, hjust = 1),  # Rotate x-axis labels
    legend.position = "none"  # Remove legend
  )
dev.off()


groupA = "Clinical (BRCA)" # "LogiCAR (N = 5)", "LogiCAR (N = 4)", "LogiCAR (N = 3)", "LogiCAR (N = 2)", "LogiCAR (N = 1)", "Kwon et al.", "Dannenfelser et al.", "Clinical (BRCA)", 
groupB = "Clinical (other tumors)" # "LogiCAR (N = 4)", "LogiCAR (N = 3)", "LogiCAR (N = 2)", "LogiCAR (N = 1)", "Kwon et al.", "Dannenfelser et al.", "Clinical (BRCA)", "Clinical (other tumors)"
wilcox.test(filtered_targets_withResult_df$Efficacy[filtered_targets_withResult_df$Group2 == groupA], filtered_targets_withResult_df$Efficacy[filtered_targets_withResult_df$Group2 == groupB])


############ Boxplot with scatter points for safety comparison
pdf_file <- paste0(fig_dir,paste0("Figure2c_safety_on_Discovery_safety90.pdf"))
fig_width <- 2.3
fig_height <- 2.7
fontSize <- 1.2
pdf(pdf_file, onefile=FALSE, width = fig_width,height=fig_height)
ggplot(filtered_targets_withResult_df, aes(x = Group2, y = Safety * 100)) +
  geom_boxplot(outlier.shape = NA, color = "black", fill = NA) +  # Boxplot without showing outliers
  geom_jitter(width = 0.2, size = 1, alpha = 1, aes(color = Group2)) +  # Scatter points with group colors
  scale_color_manual(values = group_colors) +  # Apply custom colors to scatter points
  geom_hline(yintercept = 90, linetype = "dashed", color = "grey") +  # Add dashed grey line at y = 90
  labs(
    title = "",
    x = "",
    y = "Safety (%)"
  ) +
  scale_y_continuous(limits = c(0, 100)) +  # Set y-axis range from 0 to 100
  theme(
    panel.grid = element_blank(), 
    panel.background = element_rect(fill = "white"),
    axis.line.y = element_line(color = "black"),
    axis.line.x = element_line(color = "black"),
    axis.ticks.y = element_line(color = "black"),
    axis.ticks.x = element_line(color = "black"),
    axis.text.y = element_text(color = "black"),
    panel.border = element_blank(), 
    axis.text.x = element_text(color = "black", angle = 60, hjust = 1),  
    legend.position = "none"
  )
dev.off()



############ scatter plot for safety (y) vs efficacy (x)
# Define custom colors for the 8 groups

# Define custom shapes for the 8 groups
my.shapes <- c(
  "LogiCAR (N = 5)" = 16,  # Solid circle
  "LogiCAR (N = 4)" = 15,  # Solid Square 
  "LogiCAR (N = 3)" = 18,  # Solid diamond
  "LogiCAR (N = 2)" = 17,  # Solid triangle
  "LogiCAR (N = 1)" = 3,  # +
  
  "Kwon et al." = 16,  # Solid circle 
  "Dannenfelser et al." = 16,  # Solid circle
  
  "Clinical (other tumors)" = 16,  # Solid circle
  "Clinical (BRCA)" = 16       # Solid circle
)


# Create new data frame for predicted values and confidence intervals
new_data <- data.frame(Efficacy = seq(min(filtered_targets_withResult_df$Efficacy), max(filtered_targets_withResult_df$Efficacy), by = 0.01))
conf_interval <- predict(lm(Safety ~ Efficacy, data = filtered_targets_withResult_df), newdata = new_data, interval = "confidence", level = 0.95)
new_data$Safety <- conf_interval[, "fit"]
new_data$ymin <- conf_interval[, "lwr"]
new_data$ymax <- conf_interval[, "upr"]

# Scale Efficacy and Safety to percentages for plotting
new_data$Efficacy <- new_data$Efficacy * 100
new_data$Safety <- new_data$Safety * 100
new_data$ymin <- new_data$ymin * 100
new_data$ymax <- new_data$ymax * 100

# Calculate Pearson correlation coefficient (PCC) and p-value
cor_test_result <- cor.test(filtered_targets_withResult_df$Efficacy * 100, filtered_targets_withResult_df$Safety * 100, method = "pearson")
pcc <- round(cor_test_result$estimate, 2)  # Round PCC to 2 decimal places
p_value <- format(cor_test_result$p.value, scientific = TRUE, digits = 2)  # Format p-value in scientific notation

# Save the plot to a PDF file
pdf_file <- paste0(fig_dir, "Figure2c_safety_vs_safety_scatter.pdf")
fig_width <- 4 * 1.1
fig_height <- 2.7 * 1.1
fontSize <- 1.2
pdf(pdf_file, onefile = FALSE, width = fig_width, height = fig_height)

# Create the plot
ggplot(filtered_targets_withResult_df, aes(x = Efficacy * 100, y = Safety * 100, color = Group2, shape = Group2)) + 
  geom_point(size = 1.5, alpha = 1) +  # Scatter points with size and transparency
  geom_hline(yintercept = 90, linetype = "dashed", color = "grey") +  # Add dashed grey line at y = 90
  geom_smooth(method = "lm", se = FALSE, color = "black", aes(group = 1)) +  # Add regression line for all data
  geom_ribbon(data = new_data, aes(x = Efficacy, ymin = ymin, ymax = ymax), fill = "black", alpha = 0.2, inherit.aes = FALSE) +  # Add confidence interval
  scale_color_manual(values = group_colors) +  # Use custom colors
  scale_shape_manual(values = my.shapes) +  # Use custom shapes
  labs(
    title = "",
    x = "Efficacy (%)",
    y = "Safety (%)",
    color = "",  # Remove legend title for color
    shape = ""   # Remove legend title for shape
  ) +
  scale_y_continuous(limits = c(50, 100)) +  # Set y-axis range from 50 to 100
  theme(
    panel.grid = element_blank(), 
    panel.background = element_rect(fill = "white"),
    axis.line.y = element_line(color = "black"),
    axis.line.x = element_line(color = "black"),
    axis.ticks.y = element_line(color = "black"),
    axis.ticks.x = element_line(color = "black"),
    axis.text.y = element_text(color = "black"),
    panel.border = element_blank(), 
    axis.text.x = element_text(color = "black", hjust = 1)  # Adjust x-axis text alignment
  ) +
  annotate("text", x = 5, y = 70, label = paste0("r = ", pcc, "\np = ", p_value), 
           hjust = 0, vjust = 1, size = 4, color = "black")  # Add PCC and p-value annotation

dev.off()


filtered_targets_withResult_df[order(-filtered_targets_withResult_df$Efficacy), ]

```


# Figure 2D-F. calculate efficacy  of N-gene CARs (N = 1,2,3,4,5) on full external data (Stefan, Simon) of tumor cells and safety on HPA
```{r}

####################### ####################### ####################### ####################### ####################### 
####################### process single cell data, extract gene expression of unique_CARtargets_measured
####################### ####################### ####################### ####################### ####################### 

#### process tumor data (in-house datasets External)
### Stefan
geneExpr_mat_Stefan_CARs = geneExpr_mat_Stefan[unique_CARtargets_measured, , drop = F]
geneExpr_mat_Stefan_CARs <- t(geneExpr_mat_Stefan_CARs)
colnames(geneExpr_mat_Stefan_CARs) <- unique_CARtargets_measured

geneExpr_mat_Stefan_CARs_negated <- (geneExpr_mat_Stefan_CARs == 0)*1
geneExpr_mat_Stefan_CARs_negated <- Matrix(geneExpr_mat_Stefan_CARs_negated, sparse = TRUE)
colnames(geneExpr_mat_Stefan_CARs_negated) = paste0(unique_CARtargets_measured, "_not")
geneExpr_mat_Stefan_CARs_negated[1:5,1:5]

geneExpr_mat_Stefan_CARs_final_matrix <- cbind(geneExpr_mat_Stefan_CARs, geneExpr_mat_Stefan_CARs_negated)
dim(geneExpr_mat_Stefan_CARs_final_matrix)
geneExpr_mat_Stefan_CARs_final_matrix[1:5,1:5]
## convert - in gene names to _
colnames(geneExpr_mat_Stefan_CARs_final_matrix) = gsub("-", "_", colnames(geneExpr_mat_Stefan_CARs_final_matrix))

### GSE246613
geneExpr_mat_GSE246613_CARs = geneExpr_mat_GSE246613[unique_CARtargets_measured, , drop = F]
geneExpr_mat_GSE246613_CARs <- t(geneExpr_mat_GSE246613_CARs)
colnames(geneExpr_mat_GSE246613_CARs) <- unique_CARtargets_measured

geneExpr_mat_GSE246613_CARs_negated <- (geneExpr_mat_GSE246613_CARs == 0)*1
geneExpr_mat_GSE246613_CARs_negated <- Matrix(geneExpr_mat_GSE246613_CARs_negated, sparse = TRUE)
colnames(geneExpr_mat_GSE246613_CARs_negated) = paste0(unique_CARtargets_measured, "_not")
geneExpr_mat_GSE246613_CARs_negated[1:5,1:5]

geneExpr_mat_GSE246613_CARs_final_matrix <- cbind(geneExpr_mat_GSE246613_CARs, geneExpr_mat_GSE246613_CARs_negated)
dim(geneExpr_mat_GSE246613_CARs_final_matrix)
geneExpr_mat_GSE246613_CARs_final_matrix[1:5,1:5]
## convert - in gene names to _
colnames(geneExpr_mat_GSE246613_CARs_final_matrix) = gsub("-", "_", colnames(geneExpr_mat_GSE246613_CARs_final_matrix))



# efficacy RNA
efficacy_res_list1 <- calcEquivalentExpression2(filtered_targets_df$LogiCARs, geneExpr_mat_Stefan_CARs_final_matrix) 
efficacy_res_list2 <- calcEquivalentExpression2(filtered_targets_df$LogiCARs, geneExpr_mat_GSE246613_CARs_final_matrix) 
names(efficacy_res_list1) <- filtered_targets_df$LogiCARs_display
names(efficacy_res_list2) <- filtered_targets_df$LogiCARs_display

# safety RNA
safety_res_list <- calcEquivalentExpression2(filtered_targets_df$LogiCARs, data_HPA_RNA_final_matrix) 
length(safety_res_list)
length(safety_res_list[[1]])
names(safety_res_list) <- filtered_targets_df$LogiCARs_display

result_df = data.frame(LogiCARs_display = as.character(),
                       Efficacy_Stefan = as.numeric(),
                       Efficacy_Simon = as.numeric(),
                       Safety = as.numeric()
                       )
# Loop through the names in safety_res_list
for (le in names(safety_res_list)) {
  # Calculate Efficacy and Safety
  Efficacy1 <- mean(efficacy_res_list1[[le]])
  Efficacy2 <- mean(efficacy_res_list2[[le]])
  Safety <- 1 - mean(safety_res_list[[le]])
  # Append the results as a new row to the data frame
  result_df <- rbind(result_df, data.frame(LogiCARs_display = le, Efficacy_Stefan = Efficacy1, Efficacy_Simon = Efficacy2, Safety = Safety))
}
# result_df = result_df[order(-result_df$Efficacy), ]
# result_df2 = result_df[order(result_df$Safety), ]
# print(result_df, width = 200)

filtered_targets_withExternalResult_df = merge(filtered_targets_df, result_df, by = "LogiCARs_display")
dim(filtered_targets_withExternalResult_df)




########################### boxplot
filtered_targets_withExternalResult_df$Group2 = filtered_targets_withExternalResult_df$Group
filtered_targets_withExternalResult_df$Group2[filtered_targets_withExternalResult_df$Group2 == "Clinical Targets"] = "Clinical (other tumors)"
filtered_targets_withExternalResult_df$Group2[filtered_targets_withExternalResult_df$LogiCARs %in% c("MUC1", "CD274", "PROM1", "CD70", "ROR1", "PSCA", "MET", "EPCAM", "EGFR", "FOLH1", "ERBB2", "MSLN", "CD276", "TACSTD2")] = "Clinical (BRCA)"
filtered_targets_withExternalResult_df$Group2[filtered_targets_withExternalResult_df$Group2 == "LogiCAR Designer"] = paste0("LogiCAR (N = ", filtered_targets_withExternalResult_df$geneNum[filtered_targets_withExternalResult_df$Group2 == "LogiCAR Designer"], ")")
filtered_targets_withExternalResult_df$Group2[filtered_targets_withExternalResult_df$Group2 == "Previous Comptutional (bulk)"] = "Dannenfelser et al."
filtered_targets_withExternalResult_df$Group2[filtered_targets_withExternalResult_df$Group2 == "Previous Comptutional (sc)"] = "Kwon et al."
#filtered_targets_withExternalResult_df$Group2 = factor(filtered_targets_withExternalResult_df$Group2, levels = c("LogiCAR (N = 5)", "LogiCAR (N = 4)", "LogiCAR (N = 3)", "LogiCAR (N = 2)", "LogiCAR (N = 1)", "Kwon et al.", "Dannenfelser et al.", "Clinical (BRCA)", "Clinical (other tumors)"))
filtered_targets_withExternalResult_df$Group2 = factor(filtered_targets_withExternalResult_df$Group2, levels = rev(c("LogiCAR (N = 5)", "LogiCAR (N = 4)", "LogiCAR (N = 3)", "LogiCAR (N = 2)", "Kwon et al.", "Dannenfelser et al.", "LogiCAR (N = 1)", "Clinical (BRCA)", "Clinical (other tumors)")))

############ Boxplot with scatter points for efficacy comparison
pdf_file <- paste0(fig_dir,paste0("Figure2d_efficacy_on_Stefan_safety90.pdf"))
fig_width <- 2.3
fig_height <- 2.7
fontSize <- 1.2
pdf(pdf_file, onefile = FALSE, width = fig_width, height = fig_height)
ggplot(filtered_targets_withExternalResult_df, aes(x = Group2, y = Efficacy_Stefan * 100)) +
  geom_boxplot(outlier.shape = NA, color = "black", fill = NA) +  # Black boxplot with no fill
  geom_jitter(width = 0.2, size = 1, alpha = 1, aes(color = Group2)) +  # Scatter points with group colors
  scale_color_manual(values = group_colors) +  # Apply custom colors to scatter points
  labs(
    title = "",
    x = "",
    y = "Efficacy (%)"
  ) +
  scale_y_continuous(limits = c(0, 100)) +  # Set y-axis range from 0 to 100
  theme(
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "white"),
    axis.line.y = element_line(color = "black"),
    axis.line.x = element_line(color = "black"),
    axis.ticks.y = element_line(color = "black"),
    axis.ticks.x = element_line(color = "black"),
    axis.text.y = element_text(color = "black"),
    panel.border = element_blank(),
    axis.text.x = element_text(color = "black", angle = 60, hjust = 1),  # Rotate x-axis labels
    legend.position = "none"  # Remove legend
  )
dev.off()

groupA = "Kwon et al." # "LogiCAR (N = 5)", "LogiCAR (N = 4)", "LogiCAR (N = 3)", "LogiCAR (N = 2)", "LogiCAR (N = 1)", "Kwon et al.", "Dannenfelser et al.", "Clinical (BRCA)", 
groupB = "LogiCAR (N = 2)" # "LogiCAR (N = 4)", "LogiCAR (N = 3)", "LogiCAR (N = 2)", "LogiCAR (N = 1)", "Kwon et al.", "Dannenfelser et al.", "Clinical (BRCA)", "Clinical (other tumors)"
wilcox.test(filtered_targets_withExternalResult_df$Efficacy_Stefan[filtered_targets_withExternalResult_df$Group2 == groupA], filtered_targets_withExternalResult_df$Efficacy_Stefan[filtered_targets_withExternalResult_df$Group2 == groupB])


############ Boxplot with scatter points for safety comparison
pdf_file <- paste0(fig_dir,paste0("Figure2d_efficacy_on_Simon_safety90.pdf"))
fig_width <- 2.3
fig_height <- 2.7
fontSize <- 1.2
pdf(pdf_file, onefile = FALSE, width = fig_width, height = fig_height)
ggplot(filtered_targets_withExternalResult_df, aes(x = Group2, y = Efficacy_Simon * 100)) +
  geom_boxplot(outlier.shape = NA, color = "black", fill = NA) +  # Black boxplot with no fill
  geom_jitter(width = 0.2, size = 1, alpha = 1, aes(color = Group2)) +  # Scatter points with group colors
  scale_color_manual(values = group_colors) +  # Apply custom colors to scatter points
  labs(
    title = "",
    x = "",
    y = "Efficacy (%)"
  ) +
  scale_y_continuous(limits = c(0, 100)) +  # Set y-axis range from 0 to 100
  theme(
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "white"),
    axis.line.y = element_line(color = "black"),
    axis.line.x = element_line(color = "black"),
    axis.ticks.y = element_line(color = "black"),
    axis.ticks.x = element_line(color = "black"),
    axis.text.y = element_text(color = "black"),
    panel.border = element_blank(),
    axis.text.x = element_text(color = "black", angle = 60, hjust = 1),  # Rotate x-axis labels
    legend.position = "none"  # Remove legend
  )
dev.off()

groupA = "LogiCAR (N = 3)" # "LogiCAR (N = 5)", "LogiCAR (N = 4)", "LogiCAR (N = 3)", "LogiCAR (N = 2)", "LogiCAR (N = 1)", "Kwon et al.", "Dannenfelser et al.", "Clinical (BRCA)", 
groupB = "Kwon et al." # "LogiCAR (N = 4)", "LogiCAR (N = 3)", "LogiCAR (N = 2)", "LogiCAR (N = 1)", "Kwon et al.", "Dannenfelser et al.", "Clinical (BRCA)", "Clinical (other tumors)"
wilcox.test(filtered_targets_withExternalResult_df$Efficacy_Simon[filtered_targets_withExternalResult_df$Group2 == groupA], filtered_targets_withExternalResult_df$Efficacy_Simon[filtered_targets_withExternalResult_df$Group2 == groupB])



############ scatter plot for safety (y) vs mean efficacy (x)
filtered_targets_withExternalResult_df$Efficacy_mean = rowMeans(filtered_targets_withExternalResult_df[c("Efficacy_Stefan", "Efficacy_Simon")])
# Define custom shapes for the 8 groups
my.shapes <- c(
  "LogiCAR (N = 5)" = 16,  # Solid circle
  "LogiCAR (N = 4)" = 15,  # Solid Square 
  "LogiCAR (N = 3)" = 18,  # Solid diamond
  "LogiCAR (N = 2)" = 17,  # Solid triangle
  "LogiCAR (N = 1)" = 3,  # +
  
  "Kwon et al." = 16,  # Solid circle 
  "Dannenfelser et al." = 16,  # Solid circle
  
  "Clinical (other tumors)" = 16,  # Solid circle
  "Clinical (BRCA)" = 16       # Solid circle
)

# Create new data frame for predicted values and confidence intervals
new_data <- data.frame(Efficacy_mean = seq(min(filtered_targets_withExternalResult_df$Efficacy_mean), max(filtered_targets_withExternalResult_df$Efficacy_mean), by = 0.01))
conf_interval <- predict(lm(Safety ~ Efficacy_mean, data = filtered_targets_withExternalResult_df), newdata = new_data, interval = "confidence", level = 0.95)
new_data$Safety <- conf_interval[, "fit"]
new_data$ymin <- conf_interval[, "lwr"]
new_data$ymax <- conf_interval[, "upr"]

# Scale Efficacy_mean and Safety to percentages for plotting
new_data$Efficacy_mean <- new_data$Efficacy_mean * 100
new_data$Safety <- new_data$Safety * 100
new_data$ymin <- new_data$ymin * 100
new_data$ymax <- new_data$ymax * 100

# Calculate Pearson correlation coefficient (PCC) and p-value
cor_test_result <- cor.test(filtered_targets_withExternalResult_df$Efficacy_mean * 100, filtered_targets_withExternalResult_df$Safety * 100, method = "pearson")
pcc <- round(cor_test_result$estimate, 2)  # Round PCC to 2 decimal places
p_value <- format(cor_test_result$p.value, scientific = TRUE, digits = 2)  # Format p-value in scientific notation

# Save the plot to a PDF file
pdf_file <- paste0(fig_dir,paste0("Figure2h_externalEfficacy_vs_safety_scatter.pdf"))
fig_width <- 4 * 1.1
fig_height <- 2.7 * 1.1
fontSize <- 1.2
pdf(pdf_file, onefile = FALSE, width = fig_width, height = fig_height)

# Create the plot
ggplot(filtered_targets_withExternalResult_df, aes(x = Efficacy_mean * 100, y = Safety * 100, color = Group2, shape = Group2)) + 
  geom_point(size = 1.5, alpha = 1) +  # Scatter points with size and transparency
  geom_hline(yintercept = 90, linetype = "dashed", color = "grey") +  # Add dashed grey line at y = 90
  geom_smooth(method = "lm", se = FALSE, color = "black", aes(group = 1)) +  # Add regression line for all data
  geom_ribbon(data = new_data, aes(x = Efficacy_mean, ymin = ymin, ymax = ymax), fill = "black", alpha = 0.2, inherit.aes = FALSE) +  # Add confidence interval
  scale_color_manual(values = group_colors) +  # Use custom colors
  scale_shape_manual(values = my.shapes) +  # Use custom shapes
  labs(
    title = "",
    x = "Efficacy (%)",
    y = "Safety (%)",
    color = "",  # Remove legend title for color
    shape = ""   # Remove legend title for shape
  ) +
  scale_y_continuous(limits = c(50, 100)) +  # Set y-axis range from 50 to 100
  theme(
    panel.grid = element_blank(), 
    panel.background = element_rect(fill = "white"),
    axis.line.y = element_line(color = "black"),
    axis.line.x = element_line(color = "black"),
    axis.ticks.y = element_line(color = "black"),
    axis.ticks.x = element_line(color = "black"),
    axis.text.y = element_text(color = "black"),
    panel.border = element_blank(), 
    axis.text.x = element_text(color = "black", hjust = 1)  # Adjust x-axis text alignment
  ) +
  annotate("text", x = 5, y = 70, label = paste0("r = ", pcc, "\np = ", p_value), 
           hjust = 0, vjust = 1, size = 4, color = "black")  # Add PCC and p-value annotation

dev.off()

filtered_targets_withExternalResult_df[order(-filtered_targets_withExternalResult_df$Efficacy_mean), ]

```

# Figure 2G. Rank BRCA CAR targets based on average efficacy on all 17 datasets
```{r}


####################### ####################### ####################### ####################### ####################### 
####################### process single cell data, extract gene expression of unique_CARtargets_measured
####################### ####################### ####################### ####################### ####################### 

#### process tumor data (all tumor cells from public datasets without sampling)
tumor_15public_mat_CARs = tumor_15public_mat[unique_CARtargets_measured, , drop = F]
tumor_15public_mat_CARs <- t(tumor_15public_mat_CARs > 0)
dim(tumor_15public_mat_CARs)
tumor_15public_mat_CARs[1:5,1:5]
tumor_15public_mat_negated <- tumor_15public_mat_CARs == 0
tumor_15public_mat_negated <- as(tumor_15public_mat_negated, "CsparseMatrix")
colnames(tumor_15public_mat_negated) = paste0(unique_CARtargets_measured, "_not")
tumor_15public_mat_negated[1:5,1:5]
tumor_15public_final_matrix <- cbind(tumor_15public_mat_CARs, tumor_15public_mat_negated)
dim(tumor_15public_final_matrix)
tumor_15public_final_matrix[1:5,1:5]
## convert - in gene names to _
colnames(tumor_15public_final_matrix) = gsub("-", "_", colnames(tumor_15public_final_matrix))


### efficacy
efficacy_res_list3 <- calcEquivalentExpression2(filtered_targets_df$LogiCARs, tumor_15public_final_matrix) 
names(efficacy_res_list3) <- filtered_targets_df$LogiCARs_display



# Initialize the result data frame with the appropriate structure
result_df <- data.frame(
  LogiCARs_display = character(), # First column as character
  matrix(numeric(), ncol = length(unique_15public_datasets), nrow = 0) # Rest columns as numeric
)
colnames(result_df) <- c("LogiCARs_display", paste0("Efficacy_", unique_15public_datasets)) # Set column names

# Loop through the names in safety_res_list
for (le in names(efficacy_res_list3)) {
  # Calculate Efficacy for each breakpoint
  Efficacy_vec <- numeric() # Initialize as numeric for better performance
  for (bp in seq_len(length(tumor_15public_break_points) - 1)) {
    start_pt <- tumor_15public_break_points[bp] + 1
    end_pt <- tumor_15public_break_points[bp + 1]
    Efficacy <- mean(efficacy_res_list3[[le]][start_pt:end_pt], na.rm = TRUE) # Avoid NA issues
    Efficacy_vec <- c(Efficacy_vec, Efficacy)
  }
  
  # Append the results as a new row to the data frame
  new_row <- data.frame(
    LogiCARs_display = le, # First column
    t(Efficacy_vec) # Transpose the vector to match the number of columns
  )
  colnames(new_row) <- colnames(result_df) # Ensure the column names match
  result_df <- rbind(result_df, new_row) # Add the new row to the result_df
}


filtered_targets_15DatasetsResult_df = merge(filtered_targets_df, result_df, by = "LogiCARs_display")
dim(filtered_targets_15DatasetsResult_df)
filtered_targets_15DatasetsResult_df$geneNum = NULL
filtered_targets_15DatasetsResult_df$Group = NULL
colnames(filtered_targets_15DatasetsResult_df)


### merge filtered_targets_15DatasetsResult_df with filtered_targets_withExternalResult_df
filtered_targets_17DatasetsResult_df = merge(filtered_targets_15DatasetsResult_df, filtered_targets_withExternalResult_df, by = "LogiCARs_display")
filtered_targets_17DatasetsResult_df$Efficacy_mean = NULL

# To replace _not at the end of words with ! as a prefix 
filtered_targets_17DatasetsResult_df$LogiCARs_display <- gsub(
  pattern = "\\b(\\w+)_not\\b", 
  replacement = "!\\1", 
  filtered_targets_17DatasetsResult_df$LogiCARs_display
)
## convert _ in gene names to -
filtered_targets_17DatasetsResult_df$LogiCARs_display = gsub("_", "-", filtered_targets_17DatasetsResult_df$LogiCARs_display)

### extract only N<=3 and safety > 0.9 targets
filtered_targets_17DatasetsResult_df_safer = filtered_targets_17DatasetsResult_df[(filtered_targets_17DatasetsResult_df$Safety >= 0.9) & (filtered_targets_17DatasetsResult_df$geneNum <= 3), ]
filtered_targets_17DatasetsResult_df_safer$Safety = NULL
filtered_targets_17DatasetsResult_df_safer$geneNum = NULL
#filtered_targets_17DatasetsResult_df_safer$Group2 = NULL

#filtered_targets_17DatasetsResult_df_safer$Group[filtered_targets_17DatasetsResult_df_safer$Group == "LogiCAR Designer"] = "LogiCAR (N \u2264 3)"
#filtered_targets_17DatasetsResult_df_safer$Group[filtered_targets_17DatasetsResult_df_safer$Group == "Previous Comptutional (bulk)"] = "Dannenfelser et al."
#filtered_targets_17DatasetsResult_df_safer$Group[filtered_targets_17DatasetsResult_df_safer$Group == "Previous Comptutional (sc)"] = "Kwon et al."
#filtered_targets_17DatasetsResult_df_safer$Group[filtered_targets_17DatasetsResult_df_safer$Group2 == "Clinical (BRCA)"] = "Clinical (BRCA)"
#filtered_targets_17DatasetsResult_df_safer$Group[filtered_targets_17DatasetsResult_df_safer$Group2 == "Clinical (other tumors)"] = "Clinical (other tumors)"
filtered_targets_17DatasetsResult_df_safer$Group <- as.character(filtered_targets_17DatasetsResult_df_safer$Group2)
unique(filtered_targets_17DatasetsResult_df_safer$Group)
unique(filtered_targets_17DatasetsResult_df_safer$Group2)


#### relabel another Kwon target that overlaps with ours (PRLR | VTCN1)
filtered_targets_17DatasetsResult_df_safer$LogiCARs_display[filtered_targets_17DatasetsResult_df_safer$LogiCARs_display == "PRLR | VTCN1"] = "VTCN1 | PRLR*"
filtered_targets_17DatasetsResult_df_safer$LogiCARs_display[filtered_targets_17DatasetsResult_df_safer$LogiCARs_display == "PRLR | ENPP5"] = "ENPP5 | PRLR"
filtered_targets_17DatasetsResult_df_safer$LogiCARs.y = NULL
colnames(filtered_targets_17DatasetsResult_df_safer)[colnames(filtered_targets_17DatasetsResult_df_safer) == "LogiCARs.x"] = "LogiCARs"
write.csv(filtered_targets_17DatasetsResult_df_safer, paste0("/Users/changt7/Documents/00.PostDocWork/CancerResearch/01.Projects/00.Collaborating_projects/08.Sanna_CART/01.Scripts/Triplet_CART_designer/output/safety90/filtered_top_logiCARs_efficacy_safety_17Datasets.csv"), row.names = F, quote = F)

filtered_targets_17DatasetsResult_df_safer$LogiCARs <- filtered_targets_17DatasetsResult_df_safer$LogiCARs_display


filtered_targets_17DatasetsResult_df_safer$LogiCARs_display[filtered_targets_17DatasetsResult_df_safer$LogiCARs_display == "PRLR | VTCN1"] = "VTCN1 | PRLR*"
filtered_targets_17DatasetsResult_df_safer$LogiCARs_display[filtered_targets_17DatasetsResult_df_safer$LogiCARs_display == "PRLR | ENPP5"] = "ENPP5 | PRLR"
filtered_targets_17DatasetsResult_df_safer$LogiCARs.y = NULL
colnames(filtered_targets_17DatasetsResult_df_safer)[colnames(filtered_targets_17DatasetsResult_df_safer) == "LogiCARs.x"] = "LogiCARs"
write.csv(filtered_targets_17DatasetsResult_df_safer, paste0("/Users/changt7/Documents/00.PostDocWork/CancerResearch/01.Projects/00.Collaborating_projects/08.Sanna_CART/01.Scripts/Triplet_CART_designer/output/safety90/filtered_top_logiCARs_efficacy_safety_17Datasets.csv"), row.names = F, quote = F)








############################ if have already run and saved results before, run from here  ############################
filtered_targets_17DatasetsResult_df_safer = read.csv( paste0("/Users/changt7/Documents/00.PostDocWork/CancerResearch/01.Projects/00.Collaborating_projects/08.Sanna_CART/01.Scripts/Triplet_CART_designer/output/safety90/filtered_top_logiCARs_efficacy_safety_17Datasets.csv"), row.names = NULL, header = T)

filtered_targets_17DatasetsResult_df_safer <- filtered_targets_17DatasetsResult_df_safer[!(filtered_targets_17DatasetsResult_df_safer$Group2 %in% c("LogiCAR (N = 5)", "LogiCAR (N = 4)")), ]
filtered_targets_17DatasetsResult_df_safer <- filtered_targets_17DatasetsResult_df_safer[filtered_targets_17DatasetsResult_df_safer$Safety >= 0.9, ]
filtered_targets_17DatasetsResult_df_safer$LogiCARs <- filtered_targets_17DatasetsResult_df_safer$LogiCARs_display


######################## plot barplot of mean efficacy of targets across 17 datasets with errorbars and scatters, 
plot_order = c("LogiCAR (N = 3)", "LogiCAR (N = 2)", "Kwon et al.", "Dannenfelser et al.", "LogiCAR (N = 1)", "Clinical (BRCA)", "Clinical (other tumors)")

# Gather all Efficacy columns into a long format
#colnames(filtered_targets_17DatasetsResult_df_safer) = gsub("Efficay", "Efficacy", colnames(filtered_targets_17DatasetsResult_df_safer))
efficacy_long <- filtered_targets_17DatasetsResult_df_safer %>%
  pivot_longer(
    cols = starts_with("Efficacy_"), 
    names_to = "Dataset", 
    values_to = "Efficacy"
  )
efficacy_long$Dataset = gsub("Efficacy_", "", efficacy_long$Dataset)
efficacy_long$Dataset[efficacy_long$Dataset == "Stefan"] = "In-house 1"
efficacy_long$Dataset[efficacy_long$Dataset == "Simon"] = "In-house 2"
unique(efficacy_long$Dataset)

# Calculate mean and standard error for each group
summary_stats <- efficacy_long %>%
  group_by(Group2, LogiCARs) %>%
  summarise(
    Mean_Efficacy = mean(Efficacy, na.rm = TRUE),
    SE_Efficacy = sd(Efficacy, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

# Order summary_stats by desired Group order and descending Mean_Efficacy within each group
summary_stats <- summary_stats %>%
  arrange(factor(Group2, levels = plot_order), desc(Mean_Efficacy))

# Create the ordered LogiCARs factor based on the arranged summary_stats
ordered_logiCARs <- unique(summary_stats$LogiCARs)

#Now that we have the order, we need to apply it to the data frame used for plotting
efficacy_long <- efficacy_long %>%
  mutate(LogiCARs = factor(LogiCARs, levels = ordered_logiCARs))

summary_stats <- summary_stats %>%
  mutate(LogiCARs = factor(LogiCARs, levels = ordered_logiCARs))



# Ensure Group is a factor with the desired levels (do this AFTER ordering LogiCARs)
efficacy_long$Group <- factor(
  efficacy_long$Group2,
  levels = plot_order # Desired order
)

# Ensure Dataset is a factor with the desired levels
efficacy_long$Dataset <- factor(
  efficacy_long$Dataset,
  levels = c("EMTAB8107", "GSE176078", "Bassez_CohortA", "Bassez_CohortB", "Wu", "GSE161529", "SRP114962", "GSE140819", "GSE148673", "GSE150660", "GSE143423", "GSE138536", "GSE118389", "GSE75688", "GSE158724", "In-house 1", "In-house 2") # Desired order
)

# Ensure Group in summary_stats is a factor with the desired levels
summary_stats$Group <- factor(
  summary_stats$Group2,
  levels = plot_order # Desired order
)

# Define custom colors for Dataset
my.colors <- c(
  "EMTAB8107" = "#117733",  # Dark green
  "GSE176078" = "#332288",  # Dark blue
  "Bassez_CohortA" = "#88CCEE",  # Sky blue
  "Bassez_CohortB" = "#CC6677",  # Coral
  "Wu" = "#DDCC77",  # Mustard
  "GSE161529" = "#44AA99",  # Teal
  "SRP114962" = "#999933",  # Olive green
  "GSE140819" = "#882255",  # Burgundy
  "GSE148673" = "#AA4499",  # Magenta
  "GSE150660" = "#6699CC",  # Muted blue
  "GSE143423" = "#117733",  # Forest green
  "GSE138536" = "#88CCEE",  # Pale blue
  "GSE118389" = "#DDCC77",  # Pale mustard
  "GSE75688" = "#CC6677",  # Salmon
  "GSE158724" = "#44AA99",  # Sea green
  "In-house 1" = "#AA4499",  # Bright magenta
  "In-house 2" = "#882255"  # Wine red
)


# Barplot with scatter points colored by Dataset
pdf_file <- paste0(fig_dir, "Figure2I_efficacy_17datasets_ranking_barplot.pdf")
fig_width <- 11
fig_height <- 4.5
fontSize <- 1.2
pdf(pdf_file, onefile = FALSE, width = fig_width, height = fig_height)

ggplot(summary_stats, aes(x = LogiCARs, y = Mean_Efficacy * 100, fill = Group)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  geom_errorbar(
    aes(
      ymin = Mean_Efficacy * 100 - SE_Efficacy * 100,
      ymax = Mean_Efficacy * 100 + SE_Efficacy * 100
    ),
    width = 0.2, position = position_dodge(width = 0.9), color = "black"
  ) +
  geom_jitter(
    data = efficacy_long,
    aes(x = LogiCARs, y = Efficacy * 100, color = Dataset),
    width = 0.2, size = 1, alpha = 1, inherit.aes = FALSE
  ) +
  scale_color_manual(values = my.colors) +  # Use custom colors for Dataset
  scale_fill_manual(values = group_colors) +  # Use custom colors for Group
  labs(
    title = "",
    x = "",
    y = "Efficacy (%)",
    color = "Dataset",  # Legend title for color
    fill = "Group"   # Legend title for fill
  ) +
  scale_y_continuous(limits = c(0, 100)) +
  theme(
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "white"),
    axis.line.y = element_line(color = "black"),
    axis.line.x = element_line(color = "black"),
    axis.ticks.y = element_line(color = "black"),
    axis.ticks.x = element_line(color = "black"),
    axis.text.y = element_text(color = "black"),
    panel.border = element_blank(),
    axis.text.x = element_text(color = "black", angle = 90, vjust = 0.5, hjust = 1),
    legend.position = c(1.2, 0.1),
    legend.text = element_text(family = "Arial"),  # Use a Unicode-compatible font: Arial sans
    plot.margin = margin(5, 200, 5, 5)  # Add space on the right for the legend
  ) +
  guides(
    color = guide_legend(override.aes = list(size = 2), ncol = 2),  # Increase legend symbol size for scatter points
    fill = guide_legend(override.aes = list(size = 0))  # Keep bar legend symbols small
  )

dev.off()


#### pair-wise Wilcoxon test
filtered_targets_17DatasetsResult_df_safer$mean_efficacy <- rowMeans(filtered_targets_17DatasetsResult_df_safer[grepl("Efficacy", colnames(filtered_targets_17DatasetsResult_df_safer))])
filtered_targets_17DatasetsResult_df_safer <- filtered_targets_17DatasetsResult_df_safer[order(-filtered_targets_17DatasetsResult_df_safer$mean_efficacy), ]

best_logiCAR3 <- filtered_targets_17DatasetsResult_df_safer[filtered_targets_17DatasetsResult_df_safer$LogiCARs_display=="(GABRP | PRLR) | VTCN1", grepl("Efficacy", colnames(filtered_targets_17DatasetsResult_df_safer))]

best_logiCAR2 <- filtered_targets_17DatasetsResult_df_safer[filtered_targets_17DatasetsResult_df_safer$LogiCARs_display=="PRLR | GABRP", grepl("Efficacy", colnames(filtered_targets_17DatasetsResult_df_safer))]

best_Kwon <- filtered_targets_17DatasetsResult_df_safer[filtered_targets_17DatasetsResult_df_safer$LogiCARs_display=="LY6K | PRLR", grepl("Efficacy", colnames(filtered_targets_17DatasetsResult_df_safer))]

best_Dannenfilser <- filtered_targets_17DatasetsResult_df_safer[filtered_targets_17DatasetsResult_df_safer$LogiCARs_display=="EPHB3 & !FZD5", grepl("Efficacy", colnames(filtered_targets_17DatasetsResult_df_safer))]

best_logiCAR1 <- filtered_targets_17DatasetsResult_df_safer[filtered_targets_17DatasetsResult_df_safer$LogiCARs_display=="ELAPOR1", grepl("Efficacy", colnames(filtered_targets_17DatasetsResult_df_safer))]

best_ClinicalBRCA <- filtered_targets_17DatasetsResult_df_safer[filtered_targets_17DatasetsResult_df_safer$LogiCARs_display=="PROM1", grepl("Efficacy", colnames(filtered_targets_17DatasetsResult_df_safer))]

best_ClinicalOthers <- filtered_targets_17DatasetsResult_df_safer[filtered_targets_17DatasetsResult_df_safer$LogiCARs_display=="TNFRSF10B", grepl("Efficacy", colnames(filtered_targets_17DatasetsResult_df_safer))]


wilcox.test(as.numeric(best_logiCAR3), as.numeric(best_logiCAR1), paired = T)

```








# (Suppl.) efficacy of clinical targets in different tumor datasets (heatmap, clustering, identify outlier datasets)
```{r}

library(pheatmap)


unique_15public_datasets
tumor_15public_break_points
efficacy_res_list

# Initialize the result data frame with appropriate dimensions
result_df <- data.frame(matrix(NA, nrow = length(names(efficacy_res_list)), ncol = length(unique_15public_datasets)))
rownames(result_df) <- names(efficacy_res_list)
colnames(result_df) <- unique_15public_datasets

# Loop through the names in efficacy_res_list
for (le in names(efficacy_res_list)) {
  # Calculate Efficacy
  Efficacy_all <- numeric(length(unique_15public_datasets))  # Preallocate for efficiency
  
  for (bp_i in 1:(length(tumor_15public_break_points) - 1)) {
    # Extract relevant data for the current dataset
    data_temp <- efficacy_res_list[[le]][(tumor_15public_break_points[bp_i] + 1):tumor_15public_break_points[bp_i + 1]]
    
    # Calculate mean efficacy for the current dataset
    Efficacy_temp <- mean(data_temp, na.rm = TRUE)
    Efficacy_all[bp_i] <- Efficacy_temp
  }
  
  # Assign calculated values to the appropriate row in result_df
  result_df[le, ] <- Efficacy_all
}

print(result_df, width = 200)  # Adjust the width as needed




####### heatmap
result_df2 = t(result_df)
output_file <- paste0(fig_dir,"heatmap_15datasets_clinicalCAR_efficacy.png")
# Open a PNG graphics device
png(output_file, width = 1200, height = 800, res = 150)  # Adjust width, height, and resolution as needed
# Generate a heatmap with row and column clustering
pheatmap(result_df2,
         scale = "none",         # No scaling; keep original values
         clustering_distance_rows = "euclidean",  # Distance metric for rows
         clustering_distance_cols = "euclidean",  # Distance metric for columns
         clustering_method = "complete",          # Clustering method
         show_rownames = TRUE,   # Show row names
         show_colnames = TRUE,   # Show column names
         color = colorRampPalette(c("blue", "white", "red"))(100))  # Color scheme

dev.off()

```


# load data
```{r}

### clinical CAR targets for solid tumors
#clinical_targets_df = read.csv(file = "/Users/changt7/Documents/00.PostDocWork/CancerResearch/01.Projects/00.Collaborating_projects/08.Sanna_CART/02.Input/Constants/solid_indications.txt", header = T, sep = "\t")
clinical_targets = c("MUC1", "CD274", "PROM1", "CD70", "ROR1", "CD80", "CD86", "TNFRSF10B", "FOLR1", "MPP2", "MUC16", "PMEL", "ROR2", "KDR", "EPHA2", "PSCA", "MET", "IL13RA2", "EPCAM", "EGFR", "FOLH1", "GPC3", "ERBB2", "MSLN", "CD276", "TM4SF1", "BSG", "CEACAM5", "ALPP", "GUCY2C", "TACSTD2")

### computational targets (sc + bulk)
# Cell Systems (bulk RNAseq)
logiCAR_CellSystems = list("AN_FAP_EPHA2" = c("FAP", "EPHA2"), "AN_FAP_MET" = c("FAP", "MET"),"AN_OR2B6_BMP2" = c("OR2B6", "BMP2"),
                          "AN_OR2B6_MET" = c("OR2B6", "MET"), "AN_SLC1A4_SSTR1" = c("SLC1A4", "SSTR1"), "AN_ADAM12_KDR" = c("ADAM12", "KDR"),
                          "AN_EPHB3_FZD5" = c("EPHB3", "FZD5"), "AN_TNFRSF18_ABHD6" = c("TNFRSF18", "ABHD6"), "AN_ADAM12_GPC3" = c("ADAM12", "GPC3"),
                          "AN_MUC1_LYVE1" = c("MUC1", "LYVE1"), "AN_SLC4A5_PTH1R" = c("SLC4A5", "PTH1R"), "AN_CNIH2_TMEM88" = c("CNIH2", "TMEM88"),
                          "AN_FLVCR1_TRHDE" = c("FLVCR1", "TRHDE"), "AN_CNIH2_PVRL3" = c("CNIH2", "NECTIN3"), "AN_EPHB3_SLC16A7" = c("EPHB3", "SLC16A7"),
                          "AN_EPCAM_SSTR1" = c("EPCAM", "SSTR1"), "AN_MUC1_SLC4A4" = c("MUC1", "SLC4A4"), "AN_TNFRSF18_SLC16A7" = c("TNFRSF18", "SLC16A7"),
                          "AN_LDLRAD3_SSTR1" = c("LDLRAD3", "SSTR1")
                    )
# NBT (scRNAseq)
logiCAR_NBT = list("A_ATP5MC1_KRT18" = c("ATP5MC1", "KRT18"), "A_ATP5MC1_KRT19" = c("ATP5MC1", "KRT19"), "A_ATP5MC1_KRT8" = c("ATP5MC1", "KRT8"), 
                    "A_ATP5MC1_SPINT2" = c("ATP5MC1", "SPINT2"), "A_ATP5MC1_PPDPF" = c("ATP5MC1", "PPDPF"), "A_KRT18_MARCKSL1" = c("KRT18", "MARCKSL1"),
                    "A_KRT19_MARCKSL1" = c("KRT19", "MARCKSL1"), "A_ATP5MC1_CD9" = c("ATP5MC1", "CD9"),
                    "O_CD24_KRT19" = c("CD24", "KRT19"), "O_KRT18_KRT19" = c("KRT18", "KRT19"), "O_KRT19_KRT8" = c("KRT19", "KRT8"),
                     "O_CD24_KRT18" = c("CD24", "KRT18"), "O_KRT19_MARCKSL1" = c("KRT19", "MARCKSL1"), "O_KRT19_SPINT2" = c("KRT19", "SPINT2"),
                     "O_KRT18_MARCKSL1" = c("KRT18", "MARCKSL1"), "O_CD24_SPINT2" = c("CD24", "SPINT2"),
                    "AN_KRT19_TSPAN8" = c("KRT19", "TSPAN8"), "AN_KRT19_MUC4" = c("KRT19", "MUC4"), "AN_KRT19_MECOM" = c("KRT19", "MECOM"),
                    "AN_KRT18_TSPAN8" = c("KRT18", "TSPAN8"), "AN_KRT18_PIGR" = c("KRT18", "PIGR"), "AN_KRT19_SLC4A4" = c("KRT19", "SLC4A4"),
                    "AN_KRT19_CCND2" = c("KRT19", "CCND2"), "AN_KRT19_MUC3A" = c("KRT19", "MUC3A")
                    )
### check which genes are NOT in the surface gene list
CellSystems_genes = unique(unlist(logiCAR_CellSystems))
NBT_genes = unique(unlist(logiCAR_NBT))
CellSystems_genes[!(CellSystems_genes %in% surface_genes)]
# "BMP2": Vesicles, Plasma membrane, Primary cilium, Primary cilium transition zone (predicted secreted); BMP2 is secreted, not a transmembrane protein.
# "ABHD6": Vesicles & Nucleoplasm (predicted membrane); Anchored to membranes; primarily an intracellular-facing enzyme.
# "CNIH2": Not available (predicted membrane); Functions mainly in AMPA receptor trafficking, not stably on surface.
# "TMEM88": Not available (predicted membrane); Single-pass transmembrane protein; limited data on surface expression. potentially a CAR target
NBT_genes[(NBT_genes %in% surface_genes)]
# "ATP5MC1"  "KRT18"    "KRT19"    "KRT8"     "PPDPF"    "MARCKSL1" "MECOM"    "CCND2": none are cell surface genes
### (viable) combinations
Comptutional_CAR_bulk = c("FAP & EPHA2_not", "FAP & MET_not", "OR2B6 & BMP2_not", "OR2B6 & MET_not", "SLC1A4 & SSTR1_not", "ADAM12 & KDR_not", "EPHB3 & FZD5_not", "TNFRSF18 & ABHD6_not", "ADAM12 & GPC3_not", "MUC1 & LYVE1_not", "SLC4A5 & PTH1R_not", "CNIH2 & TMEM88_not", "FLVCR1 & TRHDE_not", "CNIH2 & NECTIN3_not", "EPHB3 & SLC16A7_not", "EPCAM & SSTR1_not", "MUC1 & SLC4A4_not", "TNFRSF18 & SLC16A7_not", "LDLRAD3 & SSTR1_not")
Comptutional_CAR_sc = c(
  "ATP5MC1 & KRT18", "ATP5MC1 & KRT19", "ATP5MC1 & KRT8", "ATP5MC1 & SPINT2", "ATP5MC1 & PPDPF",
  "KRT18 & MARCKSL1", "KRT19 & MARCKSL1", "ATP5MC1 & CD9",
  "CD24 | KRT19", "KRT18 | KRT19", "KRT19 | KRT8", "CD24 | KRT18", "KRT19 | MARCKSL1", "KRT19 | SPINT2",
  "KRT18 | MARCKSL1", "CD24 | SPINT2",
  "KRT19 & TSPAN8_not", "KRT19 & MUC4_not", "KRT19 & MECOM_not", "KRT18 & TSPAN8_not", "KRT18 & PIGR_not",
  "KRT19 & SLC4A4_not", "KRT19 & CCND2_not", "KRT19 & MUC3A_not"
)

### newly identified targets
logiCAR_combos = c('PTPRK | KIAA1324 | CA12',
                   'GABRP | KIAA1324 | TACSTD2',
                   'GABRP | KIAA1324 | PRLR', # safety >0.9
                   #'(IRAG1_not & PTPRF) | KIAA1324', 
                   '((SLC39A6 | GABRP) & ACKR4_not) | (KIAA1324 | BAMBI)'
                   )  #  # OA_O_O: ((G1 | G2) & G3) | (G1 | G2)

### merged tumor scRNAseq
seu_integrated = readRDS(file = "/Users/changt7/Documents/00.PostDocWork/CancerResearch/01.Projects/00.Collaborating_projects/08.Sanna_CART/02.Input/seu_integrated_allBRCAdatasets.rds")
colnames(seu_integrated@meta.data)

# seu_test = readRDS(file = "/Users/changt7/Documents/00.PostDocWork/CancerResearch/01.Projects/00.Collaborating_projects/08.Sanna_CART/02.Input/GSE161529/seu_Tissue.rds")
# colnames(seu_test@meta.data)
# table(seu_test$cellType)
# unique(seu_test$patient_id)

### HPA scRNAseq
seu_HPA = readRDS(file = "/Users/changt7/Documents/00.PostDocWork/CancerResearch/01.Projects/00.Collaborating_projects/08.Sanna_CART/02.Input/HPA/version24_20241022/seu_HPA_v24.rds")
colnames(seu_HPA@meta.data)
unique(seu_HPA$Tissue_grouped)
unique(seu_HPA$cell_group_custom)

### HPA protein-level expression



```


# efficacy landscape of clinical targets (for all solid tumors), computational targets (sc + bulk), and newly identified targets
```{r}

# extract gene expression counts to table / data frame (whichever is faster / more efficient)
seu_integrated
setdiff(surface_genes, rownames(seu_integrated))

# Step 2: Extract the gene expression matrix (counts) for the subsetted object
tumorCells_surfaceGenes_counts <- GetAssayData(seu_integrated, assay = "RNA", slot = "counts")
colnames(tumorCells_surfaceGenes_counts) = NULL

# Step 3: Convert the gene expression matrix to a binary matrix
# (0 = FALSE, >0 = TRUE)
tumorCells_surfaceGenes_binaryExpr <- tumorCells_surfaceGenes_counts > 0
colnames(tumorCells_surfaceGenes_binaryExpr)[1:10]
rownames(tumorCells_surfaceGenes_binaryExpr)[1:10]
tumorCells_surfaceGenes_negated <- tumorCells_surfaceGenes_counts == 0
tumorCells_surfaceGenes_negated_sparse <- as(tumorCells_surfaceGenes_negated, "CsparseMatrix")
rownames(tumorCells_surfaceGenes_negated) = paste0(rownames(tumorCells_surfaceGenes_negated), "_not")
rownames(tumorCells_surfaceGenes_negated_sparse) = paste0(rownames(tumorCells_surfaceGenes_negated_sparse), "_not")

combined_matrix <- rbind(tumorCells_surfaceGenes_binaryExpr, tumorCells_surfaceGenes_negated)
combined_matrix2 <- rbind(tumorCells_surfaceGenes_binaryExpr, tumorCells_surfaceGenes_negated_sparse)
class(combined_matrix)
class(combined_matrix2)
combined_matrix = t(combined_matrix)
combined_matrix2 = t(combined_matrix2)

# clinical_targets
# Comptutional_CAR_bulk
# Comptutional_CAR_sc
# logiCAR_combos

clinical_targets_expr = combined_matrix[, clinical_targets]
# Measure time with system.time()
timing <- system.time({
  Comptutional_CAR_bulk_expr1 <- calcEquivalentExpression(Comptutional_CAR_bulk, combined_matrix)
})
# Print timing results
print(timing)

timing2 <- system.time({
  Comptutional_CAR_bulk_expr2 <- calcEquivalentExpression2(Comptutional_CAR_bulk, combined_matrix)
})
# Print timing results
print(timing2)


timing3 <- system.time({
  Comptutional_CAR_bulk_expr3 <- calcEquivalentExpression(Comptutional_CAR_bulk, combined_matrix2)
})
# Print timing results
print(timing3)

#subset_matrix <- combined_matrix2[1:5000, , drop = FALSE]
logiCAR_combos = c('PTPRK | KIAA1324 | CA12',
                   'GABRP | KIAA1324 | TACSTD2',
                   'GABRP | KIAA1324 | PRLR', # safety >0.9
                   #'(IRAG1_not & PTPRF) | KIAA1324', 
                   '((SLC39A6 | GABRP) & ACKR4_not) | (KIAA1324 | BAMBI)',
                   "CD24 | SPINT2",
                    "FAP & EPHA2_not", "FAP & MET_not",  
                    "OR2B6 & MET_not", "SLC1A4 & SSTR1_not", "ADAM12 & KDR_not", 
                    "EPHB3 & FZD5_not", "ADAM12 & GPC3_not", 
                    "MUC1 & LYVE1_not", "SLC4A5 & PTH1R_not", 
                    "FLVCR1 & TRHDE_not", "EPHB3 & SLC16A7_not", 
                    "EPCAM & SSTR1_not", "MUC1 & SLC4A4_not", "TNFRSF18 & SLC16A7_not", 
                    "LDLRAD3 & SSTR1_not"
                   )
timing4 <- system.time({
  Comptutional_CAR_bulk_expr4 <- calcEquivalentExpression2(logiCAR_combos, combined_matrix2)
  #Comptutional_CAR_bulk_expr4 <- calcEquivalentExpression2(logiCAR_combos, subset_matrix)
})
# Print timing results
print(timing4)
means_list <- lapply(Comptutional_CAR_bulk_expr4, mean)
sorted_list <- means_list[order(unlist(means_list))]
print(sorted_list)

```

# safety landscape of clinical targets (for all solid tumors), computational targets (sc + bulk), and newly identified targets
```{r}

```


#  (not used!!) calculate efficacy and safety of N-gene CARs (N = 1,2,3,4,5) on full data of tumor cells and HPA normal cells (use all cells)
```{r}

####################### ####################### ####################### ####################### ####################### 
####################### process single cell data, extract gene expression of unique_CARtargets_measured
####################### ####################### ####################### ####################### ####################### 



#### process tumor data (in-house datasets External)
### Stefan
geneExpr_mat_Stefan_CARs = geneExpr_mat_Stefan[unique_CARtargets_measured, , drop = F]
geneExpr_mat_Stefan_CARs <- t(geneExpr_mat_Stefan_CARs)
colnames(geneExpr_mat_Stefan_CARs) <- unique_CARtargets_measured

geneExpr_mat_Stefan_CARs_negated <- (geneExpr_mat_Stefan_CARs == 0)*1
geneExpr_mat_Stefan_CARs_negated <- Matrix(geneExpr_mat_Stefan_CARs_negated, sparse = TRUE)
colnames(geneExpr_mat_Stefan_CARs_negated) = paste0(unique_CARtargets_measured, "_not")
geneExpr_mat_Stefan_CARs_negated[1:5,1:5]

geneExpr_mat_Stefan_CARs_final_matrix <- cbind(geneExpr_mat_Stefan_CARs, geneExpr_mat_Stefan_CARs_negated)
dim(geneExpr_mat_Stefan_CARs_final_matrix)
geneExpr_mat_Stefan_CARs_final_matrix[1:5,1:5]
## convert - in gene names to _
colnames(geneExpr_mat_Stefan_CARs_final_matrix) = gsub("-", "_", colnames(geneExpr_mat_Stefan_CARs_final_matrix))

### GSE246613
geneExpr_mat_GSE246613_CARs = geneExpr_mat_GSE246613[unique_CARtargets_measured, , drop = F]
geneExpr_mat_GSE246613_CARs <- t(geneExpr_mat_GSE246613_CARs)
colnames(geneExpr_mat_GSE246613_CARs) <- unique_CARtargets_measured

geneExpr_mat_GSE246613_CARs_negated <- (geneExpr_mat_GSE246613_CARs == 0)*1
geneExpr_mat_GSE246613_CARs_negated <- Matrix(geneExpr_mat_GSE246613_CARs_negated, sparse = TRUE)
colnames(geneExpr_mat_GSE246613_CARs_negated) = paste0(unique_CARtargets_measured, "_not")
geneExpr_mat_GSE246613_CARs_negated[1:5,1:5]

geneExpr_mat_GSE246613_CARs_final_matrix <- cbind(geneExpr_mat_GSE246613_CARs, geneExpr_mat_GSE246613_CARs_negated)
dim(geneExpr_mat_GSE246613_CARs_final_matrix)
geneExpr_mat_GSE246613_CARs_final_matrix[1:5,1:5]
## convert - in gene names to _
colnames(geneExpr_mat_GSE246613_CARs_final_matrix) = gsub("-", "_", colnames(geneExpr_mat_GSE246613_CARs_final_matrix))


############ process HPA RNA data
data_HPA_df_onlyData = seu_HPA_RNA_counts_matrix[unique_CARtargets_measured, , drop = F]
data_HPA_df_onlyData <- t(data_HPA_df_onlyData > 0)
colnames(data_HPA_df_onlyData) <- unique_CARtargets_measured

data_HPA_df_onlyData_negated <- data_HPA_df_onlyData == 0
data_HPA_df_onlyData_negated <- as(data_HPA_df_onlyData_negated, "CsparseMatrix")
colnames(data_HPA_df_onlyData_negated) = paste0(unique_CARtargets_measured, "_not")

data_HPA_final_matrix <- cbind(data_HPA_df_onlyData, data_HPA_df_onlyData_negated)
dim(data_HPA_final_matrix)
data_HPA_final_matrix[1:5,1:5]
## convert - in gene names to _
colnames(data_HPA_final_matrix) = gsub("-", "_", colnames(data_HPA_final_matrix))
max(data_HPA_final_matrix)
dim(data_HPA_final_matrix)


# efficacy RNA
efficacy_res_list <- calcEquivalentExpression2(filtered_targets_vec, geneExpr_mat_GSE246613_CARs_final_matrix) 
# first para: targets_vec unique_genes_in_HPAprotein filtered_targets_vec unique_clinical_targets_measured
# second para: tumor_15public_final_matrix geneExpr_mat_GSE246613_CARs_final_matrix geneExpr_mat_Stefan_CARs_final_matrix
length(efficacy_res_list)
length(efficacy_res_list[[1]])

# safety RNA
safety_res_list <- calcEquivalentExpression2(filtered_targets_vec, data_HPA_final_matrix) # targets_vec unique_genes_in_HPAprotein filtered_targets_vec
length(safety_res_list)
length(safety_res_list[[1]])

result_df = data.frame(Efficay = as.numeric(),
                       Safety = as.numeric()
                       )
# Loop through the names in safety_res_list
for (le in names(safety_res_list)) {
  # Calculate Efficacy and Safety
  Efficacy <- mean(efficacy_res_list[[le]])
  Safety <- 1 - mean(safety_res_list[[le]])
  # Append the results as a new row to the data frame
  result_df <- rbind(result_df, data.frame(Efficacy = Efficacy, Safety = Safety))
}
rownames(result_df) = names(safety_res_list)
result_df = result_df[order(-result_df$Efficacy), ]
result_df2 = result_df[order(result_df$Safety), ]
print(result_df, width = 200)



```


# (not used!!) Figure 2-d. calculate efficacy and safety of N-gene CARs (N = 1,2,3,4,5) on full external data (Stefan, Simon) of tumor cells and T.S. normal cells (use all cells)
```{r}

####################### ####################### ####################### ####################### ####################### 
####################### process single cell data, extract gene expression of unique_CARtargets_measured
####################### ####################### ####################### ####################### ####################### 

#### process tumor data (public datasets Discovery)
tumor_15public_mat <- t(tumor_15public_mat > 0)
dim(tumor_15public_mat)
tumor_15public_mat[1:5,1:5]
tumor_15public_mat_negated <- tumor_15public_mat == 0
tumor_15public_mat_negated <- as(tumor_15public_mat_negated, "CsparseMatrix")
colnames(tumor_15public_mat_negated) = paste0(geneNAs, "_not")
tumor_15public_mat_negated[1:5,1:5]
tumor_15public_final_matrix <- cbind(tumor_15public_mat, tumor_15public_mat_negated)
dim(tumor_15public_final_matrix)
tumor_15public_final_matrix[1:5,1:5]
## convert - in gene names to _
colnames(tumor_15public_final_matrix) = gsub("-", "_", colnames(tumor_15public_final_matrix))


#### process tumor data (in-house datasets External)
### Stefan
geneExpr_mat_Stefan_CARs = geneExpr_mat_Stefan[unique_CARtargets_measured, , drop = F]
geneExpr_mat_Stefan_CARs <- t(geneExpr_mat_Stefan_CARs)
colnames(geneExpr_mat_Stefan_CARs) <- unique_CARtargets_measured

geneExpr_mat_Stefan_CARs_negated <- (geneExpr_mat_Stefan_CARs == 0)*1
geneExpr_mat_Stefan_CARs_negated <- Matrix(geneExpr_mat_Stefan_CARs_negated, sparse = TRUE)
colnames(geneExpr_mat_Stefan_CARs_negated) = paste0(unique_CARtargets_measured, "_not")
geneExpr_mat_Stefan_CARs_negated[1:5,1:5]

geneExpr_mat_Stefan_CARs_final_matrix <- cbind(geneExpr_mat_Stefan_CARs, geneExpr_mat_Stefan_CARs_negated)
dim(geneExpr_mat_Stefan_CARs_final_matrix)
geneExpr_mat_Stefan_CARs_final_matrix[1:5,1:5]
## convert - in gene names to _
colnames(geneExpr_mat_Stefan_CARs_final_matrix) = gsub("-", "_", colnames(geneExpr_mat_Stefan_CARs_final_matrix))

### GSE246613
geneExpr_mat_GSE246613_CARs = geneExpr_mat_GSE246613[unique_CARtargets_measured, , drop = F]
geneExpr_mat_GSE246613_CARs <- t(geneExpr_mat_GSE246613_CARs)
colnames(geneExpr_mat_GSE246613_CARs) <- unique_CARtargets_measured

geneExpr_mat_GSE246613_CARs_negated <- (geneExpr_mat_GSE246613_CARs == 0)*1
geneExpr_mat_GSE246613_CARs_negated <- Matrix(geneExpr_mat_GSE246613_CARs_negated, sparse = TRUE)
colnames(geneExpr_mat_GSE246613_CARs_negated) = paste0(unique_CARtargets_measured, "_not")
geneExpr_mat_GSE246613_CARs_negated[1:5,1:5]

geneExpr_mat_GSE246613_CARs_final_matrix <- cbind(geneExpr_mat_GSE246613_CARs, geneExpr_mat_GSE246613_CARs_negated)
dim(geneExpr_mat_GSE246613_CARs_final_matrix)
geneExpr_mat_GSE246613_CARs_final_matrix[1:5,1:5]
## convert - in gene names to _
colnames(geneExpr_mat_GSE246613_CARs_final_matrix) = gsub("-", "_", colnames(geneExpr_mat_GSE246613_CARs_final_matrix))


############ process HPA RNA data
data_HPA_df_onlyData = seu_HPA_RNA_counts_matrix[unique_CARtargets_measured, , drop = F]
data_HPA_df_onlyData <- t(data_HPA_df_onlyData > 0)
colnames(data_HPA_df_onlyData) <- unique_CARtargets_measured

data_HPA_df_onlyData_negated <- data_HPA_df_onlyData == 0
data_HPA_df_onlyData_negated <- as(data_HPA_df_onlyData_negated, "CsparseMatrix")
colnames(data_HPA_df_onlyData_negated) = paste0(unique_CARtargets_measured, "_not")

data_HPA_final_matrix <- cbind(data_HPA_df_onlyData, data_HPA_df_onlyData_negated)
dim(data_HPA_final_matrix)
data_HPA_final_matrix[1:5,1:5]
## convert - in gene names to _
colnames(data_HPA_final_matrix) = gsub("-", "_", colnames(data_HPA_final_matrix))
max(data_HPA_final_matrix)
dim(data_HPA_final_matrix)


# efficacy RNA
efficacy_res_list <- calcEquivalentExpression2(filtered_targets_vec, geneExpr_mat_GSE246613_CARs_final_matrix) 
# first para: targets_vec unique_genes_in_HPAprotein filtered_targets_vec unique_clinical_targets_measured
# second para: tumor_15public_final_matrix geneExpr_mat_GSE246613_CARs_final_matrix geneExpr_mat_Stefan_CARs_final_matrix
length(efficacy_res_list)
length(efficacy_res_list[[1]])

# safety RNA
safety_res_list <- calcEquivalentExpression2(filtered_targets_vec, data_HPA_final_matrix) # targets_vec unique_genes_in_HPAprotein filtered_targets_vec
length(safety_res_list)
length(safety_res_list[[1]])

result_df = data.frame(Efficay = as.numeric(),
                       Safety = as.numeric()
                       )
# Loop through the names in safety_res_list
for (le in names(safety_res_list)) {
  # Calculate Efficacy and Safety
  Efficacy <- mean(efficacy_res_list[[le]])
  Safety <- 1 - mean(safety_res_list[[le]])
  # Append the results as a new row to the data frame
  result_df <- rbind(result_df, data.frame(Efficacy = Efficacy, Safety = Safety))
}
rownames(result_df) = names(safety_res_list)
result_df = result_df[order(-result_df$Efficacy), ]
result_df2 = result_df[order(result_df$Safety), ]
print(result_df, width = 200)



```
