---
title: "Comprehensive evaluation of the safety of potential CARs and logic-gated CARs"
author: "Tiangen Chang tiangen.chang@nih.gov"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---



# load required package
```{r, include=FALSE}

library(SeuratData)
library(SeuratDisk)
library(caret)
library(Seurat)
library(dplyr)
library(ggplot2)
library(dittoSeq)
library(SeuratWrappers)
library(patchwork)
library(RColorBrewer)
library(stringr)
library(tidyr)
library(data.table)
library(Matrix)

```

# enables R to show special characters like ≤
```{r}

library(showtext) # this package enables R to show special characters like ≤
# Add a font and enable `showtext`
font_add(family = "Arial", regular = "path/to/arial.ttf")  # Replace with the path to your Arial font file
showtext_auto()

```

# set input and output directories & parameters
```{r}

data_dir <- paste0('/02.Input/')
processed_dir <- paste0('/03.Results/')
fig_dir = paste0('/03.Results/Figures/')
surface_genes = read.csv(file = "/02.Input/Constants/cell_surface_genes_wgene_and_clinical_uniqueStandardNames.txt", header = T) 
surface_genes = as.character(surface_genes[,1])

# Define custom colors for each group
group_colors <- c(
  "LogiCAR (N = 5)" = "#08306b",  # Deep blue
  "LogiCAR (N = 4)" = "#2171b5",  # Dark blue
  "LogiCAR (N = 3)" = "#4292c6",  # Medium blue
  "LogiCAR (N = 2)" = "#6baed6",  # Light blue
  "LogiCAR (N = 1)" = "#9ecae1",  # Very light blue
  "Kwon et al." = "#006d2c",      # Deep green
  "Dannenfelser et al." = "#41ab5d",  # Light green
  "Clinical (BRCA)" = "#cb181d",  # Deep red
  "Clinical (other tumors)" = "#fb6a4a"  # Light red (pink)
)

```

# load common functions
```{r}

source("00.Common_Functions.R")

```


# load and process input data (CAR targets, HPA scRNAseq seurat, HPA protein expression matrix)
```{r}

# local
HPA_data_dir_RNA = "/02.Input/HPA/version24_20241022/rna/"
HPA_data_dir_protein = "/02.Input/HPA/version24_20241022/protein/"

###### load candidate CARs and logic-gated CARs
candidate_CARs_df = read.csv(paste0("/01.Scripts/Triplet_CART_designer/output/safety90/filtered_top_logiCARs_efficacy_safety_17Datasets.csv"))
### select N<=3 candidate CARs
candidate_CARs_df = candidate_CARs_df[candidate_CARs_df$geneNum <=3, ]

### rank first by group2, then by efficacy
plot_order = rev(c("LogiCAR (N = 3)", "LogiCAR (N = 2)", "Kwon et al.", "Dannenfelser et al.", "LogiCAR (N = 1)", "Clinical (BRCA)", "Clinical (other tumors)"))
candidate_CARs_df$Group2 = factor(candidate_CARs_df$Group2, levels = plot_order)
candidate_CARs_df$Efficacy_mean = rowMeans(candidate_CARs_df[grepl("Effic", colnames(candidate_CARs_df))])
candidate_CARs_df <- candidate_CARs_df[order(candidate_CARs_df$Group2, -candidate_CARs_df$Efficacy_mean), ]

candidate_CARs = candidate_CARs_df$LogiCARs
candidate_CARs_display = candidate_CARs_df$LogiCARs_display
cleaned_combos <- gsub("[|&()]", "", candidate_CARs)  # Remove |, &, (, )
cleaned_combos <- gsub("_not", "", cleaned_combos)    # Remove _not
all_genes <- unlist(strsplit(cleaned_combos, "\\s+"))
unique_CARtargets_measured <- unique(all_genes)


###### load HPA protein data
load(paste0(HPA_data_dir_protein,"HPA_protein_expression_matrix_Tissues_withUncertainProteins.rdata")) # var name: expression_matrix_Tissues
load(paste0(HPA_data_dir_protein,"HPA_protein_expression_matrix_CellGroups_withUncertainProteins.rdata")) # var name: expression_matrix_Cells
HPA_Protein_matrix_Tissues <- expression_matrix_Tissues
## convert - in gene names to _
rownames(HPA_Protein_matrix_Tissues) <- gsub("-", "_", rownames(HPA_Protein_matrix_Tissues))
HPA_Protein_matrix_Cells <- expression_matrix_Cells
## convert - in gene names to _
rownames(HPA_Protein_matrix_Cells) <- gsub("-", "_", rownames(HPA_Protein_matrix_Cells))
rm(expression_matrix_Tissues, expression_matrix_Cells)
allGenes_HPA_protein = intersect(rownames(HPA_Protein_matrix_Tissues), rownames(HPA_Protein_matrix_Cells)) 

### process HPA protein expression data
## cell level protein expression matrix HPA_Protein_matrix_Cells
HPA_Protein_matrix_Cells_filled <- matrix(
  NA, 
  nrow = length(unique_CARtargets_measured), 
  ncol = ncol(HPA_Protein_matrix_Cells),
  dimnames = list(unique_CARtargets_measured, colnames(HPA_Protein_matrix_Cells))
)
# Fill in the existing data
common_rows <- intersect(rownames(HPA_Protein_matrix_Cells), unique_CARtargets_measured)
HPA_Protein_matrix_Cells_filled[common_rows, ] <- HPA_Protein_matrix_Cells[common_rows, ]
# Resulting matrix will include all rows from unique_CARtargets_measured, with NAs for missing rows
HPA_Protein_matrix_Cells <- HPA_Protein_matrix_Cells_filled
HPA_Protein_matrix_Cells <- t(HPA_Protein_matrix_Cells)

HPA_Protein_matrix_Cells_negated <- 3 - HPA_Protein_matrix_Cells # use max value to subtract 
colnames(HPA_Protein_matrix_Cells_negated) = paste0(colnames(HPA_Protein_matrix_Cells_negated), "_not")

HPA_Protein_matrix_Cells_final_matrix <- cbind(HPA_Protein_matrix_Cells, HPA_Protein_matrix_Cells_negated)
dim(HPA_Protein_matrix_Cells_final_matrix)
HPA_Protein_matrix_Cells_final_matrix[1:5,1:5]
# Remove the row named "Visual"
HPA_Protein_matrix_Cells_final_matrix <- HPA_Protein_matrix_Cells_final_matrix[
  rownames(HPA_Protein_matrix_Cells_final_matrix) != "Visual", 
  ]

## tissue level protein expression matrix HPA_Protein_matrix_Tissues
HPA_Protein_matrix_Tissues_filled <- matrix(
  NA, 
  nrow = length(unique_CARtargets_measured), 
  ncol = ncol(HPA_Protein_matrix_Tissues),
  dimnames = list(unique_CARtargets_measured, colnames(HPA_Protein_matrix_Tissues))
)
# Fill in the existing data
common_rows <- intersect(rownames(HPA_Protein_matrix_Tissues), unique_CARtargets_measured)
HPA_Protein_matrix_Tissues_filled[common_rows, ] <- HPA_Protein_matrix_Tissues[common_rows, ]
# Resulting matrix will include all rows from unique_CARtargets_measured, with NAs for missing rows
HPA_Protein_matrix_Tissues <- HPA_Protein_matrix_Tissues_filled
HPA_Protein_matrix_Tissues <- t(HPA_Protein_matrix_Tissues)

HPA_Protein_matrix_Tissues_negated <- 3 - HPA_Protein_matrix_Tissues # use max value to subtract 
colnames(HPA_Protein_matrix_Tissues_negated) = paste0(colnames(HPA_Protein_matrix_Tissues_negated), "_not")

HPA_Protein_matrix_Tissues_final_matrix <- cbind(HPA_Protein_matrix_Tissues, HPA_Protein_matrix_Tissues_negated)
dim(HPA_Protein_matrix_Tissues_final_matrix)
HPA_Protein_matrix_Tissues_final_matrix[1:5,1:5]
# Remove the row named "Visual"
HPA_Protein_matrix_Tissues_final_matrix <- HPA_Protein_matrix_Tissues_final_matrix[
  rownames(HPA_Protein_matrix_Tissues_final_matrix) != "Visual", 
  ]


### load HPA RNA data
seu_HPA_RNA = readRDS(file = paste0(paste0(HPA_data_dir_RNA,"seu_HPA_v24.rds")))  # 20025 features across 689601 samples within 1 assay 
seu_HPA_RNA <- NormalizeData(seu_HPA_RNA) # log(x/sum(x)*10000+1)

genes_HPA_RNA = rownames(seu_HPA_RNA)
seu_HPA_RNA_counts_matrix <- GetAssayData(seu_HPA_RNA, layer = "data")
data_HPA_df_onlyData = seu_HPA_RNA_counts_matrix[unique_CARtargets_measured, , drop = F]
data_HPA_df_onlyData <- t(data_HPA_df_onlyData)

# Calculate the column-wise maximums
colMax <- apply(data_HPA_df_onlyData, 2, max, na.rm = TRUE)
# Subtract the original matrix values from the column-wise maximums
data_HPA_df_onlyData_negated <- sweep(data_HPA_df_onlyData, 2, colMax, FUN = function(x, y) y - x)
colnames(data_HPA_df_onlyData_negated) = paste0(colnames(data_HPA_df_onlyData_negated), "_not")

data_HPA_final_matrix <- cbind(data_HPA_df_onlyData, data_HPA_df_onlyData_negated)
## convert - in gene names to _
colnames(data_HPA_final_matrix) = gsub("-", "_", colnames(data_HPA_final_matrix))

```

# Figure 3A (left). dotplot of RNA expression versus tissues or cell types
```{r}

candidateCARs_equivalent_RNA_Expression_res_list <- calcEquivalentExpression2(candidate_CARs, data_HPA_final_matrix) 

### Convert the list to a data frame
candidateCARs_equivalent_RNA_Expression_res_df <- as.data.frame(candidateCARs_equivalent_RNA_Expression_res_list)
colnames(candidateCARs_equivalent_RNA_Expression_res_df) <- names(candidateCARs_equivalent_RNA_Expression_res_list)
candidateCARs_equivalent_RNA_Expression_res_df = candidateCARs_equivalent_RNA_Expression_res_df[candidate_CARs]
colnames(candidateCARs_equivalent_RNA_Expression_res_df) <- candidate_CARs_display

candidateCARs_equivalent_RNA_Expression_res_df$Tissue_grouped = seu_HPA_RNA$Tissue_grouped
candidateCARs_equivalent_RNA_Expression_res_df$cell_group_custom = as.character(seu_HPA_RNA$cell_group_custom)

candidateCARs_equivalent_RNA_Expression_res_df = candidateCARs_equivalent_RNA_Expression_res_df[!(candidateCARs_equivalent_RNA_Expression_res_df$cell_group_custom %in% c("Unspecified", "Mixed")), ] # from 689601 -> 671927
candidateCARs_equivalent_RNA_Expression_res_df = candidateCARs_equivalent_RNA_Expression_res_df[!(candidateCARs_equivalent_RNA_Expression_res_df$Tissue_grouped %in% c("Unspecified")), ] # from 671927 -> 671927

save(candidateCARs_equivalent_RNA_Expression_res_df, file = "/02.Input/HPA/version24_20241022/rna/candidateCARs_77_equivalent_RNA_Expression_res_df_v2.rdata")


cells_reordered = c(
  "Heart muscle", "Neuronal", "Liver epithelial", "Alveolar", "Urinary epithelial", "Hematopoietic", "Endocrine", "Integumentary", "Endothelial/vascular", 
  "Gastrointestinal epithelial", "Immune", 
  "Female reproductive", "Male reproductive", 
  "Visual", "Skeletal/smooth muscle", "General/glandular epithelial", "Glial", "Connective/stromal", "Undifferentiated")

tissues_reordered = c(
  "Heart", "Brain", "Hepatobiliary", "Lung", "Urinary", "Pancreas", "Bone marrow", "Integumentary", "Vascular",
  "Gastrointestinal", "Lymphoid", "Female reproductive", "Male reproductive", 
  "Visual", "Skeletal/smooth muscle", "Oral", "Conducting airways", "Connective")

# Gather the data into long format
genes <- rev(candidate_CARs_display)
dotplot_data <- candidateCARs_equivalent_RNA_Expression_res_df %>%
  pivot_longer(
    cols = all_of(genes),
    names_to = "Gene",
    values_to = "Expression"
  )


################## plot tissue RNA expression dotplot ################## 
# Calculate required metrics: zero ratio and mean expression for each Tissue-Gene combination
plot_data <- dotplot_data %>%
  group_by(Tissue_grouped, Gene) %>%
  summarise(
    ExprRatio = mean(Expression > 0, na.rm = TRUE), # Proportion of zeros
    MeanExpression = mean(Expression, na.rm = TRUE), # Average expression
    .groups = "drop"
  )

plot_data <- plot_data %>%
  mutate(
    Tissue_grouped = factor(Tissue_grouped, levels = tissues_reordered),
    Gene = factor(Gene, levels = genes) # Keep genes in column order
  )

# Define a vector of colors for y-axis tick labels
group_sizes <- rev(as.numeric(table(candidate_CARs_df$Group2)))
group_names <- rev(names(table(candidate_CARs_df$Group2)))
y_tick_colors <- rep(group_colors[group_names], times = group_sizes)

# Create the plot
pdf_file <- paste0(fig_dir,paste0("Figure3A_safety_landscape_RNA_tissue.pdf"))
fig_width = 5.7
fig_height = 12.5
fontSize = 1.2
pdf(pdf_file, onefile=FALSE, width = fig_width,height=fig_height)
ggplot(plot_data, aes(x = Tissue_grouped, y = Gene)) +
  geom_point(aes(size = ExprRatio, color = MeanExpression)) + # Dot size and color
  scale_color_gradientn(
    colors = c("#440154", "#31688E", "#35B779", "#FDE725"), 
    values = scales::rescale(c(0, 1, 2, 3)), # Define the positions for the colors
    limits = c(0, 3), # Set the range of the gradient
    oob = scales::squish, # Handle out-of-bound values
    labels = c("No", "Low", "Medium", "High") # Custom labels for the ticks
  ) +
  scale_size_continuous(
    range = c(1, 5), # Control dot size range
    limits = c(0, 1) # Set the range of values for the legend
  ) +
  labs(
    x = "",
    y = "",
    size = "Expressed\nratio",
    color = "Mean\nexpression"
  ) +
  theme(panel.grid = element_blank(), 
        panel.background = element_rect(fill = "white"),
        axis.line.y = element_line(color="black"),
        axis.line.x = element_line(color="black"),
        axis.ticks.y = element_line(color="black"),
        axis.ticks.x = element_line(color="black"),
        panel.border = element_blank(), 
        axis.text.y = element_text(color = y_tick_colors),  # Assign colors to y-axis tick labels
        axis.text.x = element_text(color = "black", angle = 90, vjust = 0.5, hjust = 1),  
        legend.position = "right" 
        )
dev.off()




################## plot cell type RNA expression dotplot ################## 
# Calculate required metrics: zero ratio and mean expression for each Tissue-Gene combination
plot_data <- dotplot_data %>%
  group_by(cell_group_custom, Gene) %>%
  summarise(
    ExprRatio = mean(Expression > 0, na.rm = TRUE), # Proportion of zeros
    MeanExpression = mean(Expression, na.rm = TRUE), # Average expression
    .groups = "drop"
  )

plot_data <- plot_data %>%
  mutate(
    cell_group_custom = factor(cell_group_custom, levels = cells_reordered),
    Gene = factor(Gene, levels = genes) # Keep genes in column order
  )

# Create the plot
pdf_file <- paste0(fig_dir,paste0("Figure3A_safety_landscape_RNA_cellType.pdf"))
fig_width = 5.7
fig_height = 12.5
fontSize = 1.2
pdf(pdf_file, onefile=FALSE, width = fig_width,height=fig_height)
ggplot(plot_data, aes(x = cell_group_custom, y = Gene)) +
  geom_point(aes(size = ExprRatio, color = MeanExpression)) + # Dot size and color
  scale_color_gradientn(
    colors = c("#440154", "#31688E", "#35B779", "#FDE725"), 
    values = scales::rescale(c(0, 1, 2, 3)), # Define the positions for the colors
    limits = c(0, 3), # Set the range of the gradient
    oob = scales::squish, # Handle out-of-bound values
    labels = c("No", "Low", "Medium", "High") # Custom labels for the ticks
  ) +
  scale_size_continuous(
    range = c(1, 5), # Control dot size range
    limits = c(0, 1) # Set the range of values for the legend
  ) +
  labs(
    x = "",
    y = "",
    size = "Expressed\nratio",
    color = "Mean\nexpression"
  ) +
  theme(panel.grid = element_blank(), 
        panel.background = element_rect(fill = "white"),
        axis.line.y = element_line(color="black"),
        axis.line.x = element_line(color="black"),
        axis.ticks.y = element_line(color="black"),
        axis.ticks.x = element_line(color="black"),
        axis.text.y = element_text(color = y_tick_colors),  # Assign colors to y-axis tick labels
        panel.border = element_blank(), 
        axis.text.x = element_text(color = "black", angle = 90, vjust = 0.5, hjust = 1),  
        legend.position = "right" 
        )
dev.off()


```


# Figure 3A (right). dotplot of protein expression versus tissues or cell types
```{r}

######################################### expression in Tissues #########################################
candidateCARs_equivalent_Protein_Expression_res_list <- calcEquivalentExpression2(candidate_CARs, HPA_Protein_matrix_Tissues_final_matrix) 

### Convert the list to a data frame
candidateCARs_equivalent_Protein_Expression_res_df_Tissues <- as.data.frame(candidateCARs_equivalent_Protein_Expression_res_list)
colnames(candidateCARs_equivalent_Protein_Expression_res_df_Tissues) <- names(candidateCARs_equivalent_Protein_Expression_res_list)
candidateCARs_equivalent_Protein_Expression_res_df_Tissues = candidateCARs_equivalent_Protein_Expression_res_df_Tissues[candidate_CARs]
colnames(candidateCARs_equivalent_Protein_Expression_res_df_Tissues) <- candidate_CARs_display

candidateCARs_equivalent_Protein_Expression_res_df_Tissues$Tissue_grouped = rownames(candidateCARs_equivalent_Protein_Expression_res_df_Tissues)
save(candidateCARs_equivalent_Protein_Expression_res_df_Tissues, file = "/02.Input/HPA/version24_20241022/rna/candidateCARs_77_equivalent_Protein_Expression_res_df_Tissues_v2.rdata")


tissues_reordered = c(
  "Heart", "Brain", "Hepatobiliary", "Lung", "Urinary", "Pancreas", "Bone marrow", "Integumentary", "Endocrine",
  "Gastrointestinal", "Lymphoid", "Female reproductive", "Male reproductive", 
  "Skeletal/smooth muscle", "Oral", "Conducting airways", "Connective")

# Gather the data into long format
genes <- rev(candidate_CARs_display)
dotplot_data <- candidateCARs_equivalent_Protein_Expression_res_df_Tissues %>%
  pivot_longer(
    cols = all_of(genes),
    names_to = "Gene",
    values_to = "Expression"
  )

################## plot tissue Protein expression dotplot ################## 
# Calculate required metrics: zero ratio and mean expression for each Tissue-Gene combination
plot_data <- dotplot_data %>%
  group_by(Tissue_grouped, Gene) %>%
  summarise(
    ExprRatio = mean(Expression > 0, na.rm = TRUE), # Proportion of zeros
    MeanExpression = mean(Expression, na.rm = TRUE), # Average expression
    .groups = "drop"
  )

plot_data <- plot_data %>%
  mutate(
    Tissue_grouped = factor(Tissue_grouped, levels = tissues_reordered),
    Gene = factor(Gene, levels = genes) # Keep genes in column order
  )

# Create the plot
pdf_file <- paste0(fig_dir, "Figure3A_safety_landscape_Protein_tissue.pdf")
fig_width = 5.7
fig_height = 12.5
fontSize <- 10 # Larger font size for axis tick labels
pdf(pdf_file, onefile = FALSE, width = fig_width, height = fig_height)
ggplot(plot_data, aes(x = Tissue_grouped, y = Gene)) +
  # Main heatmap with MeanExpression
  geom_tile(aes(fill = MeanExpression), width = 1, height = 1) +
  # Add a cross for NA values
  geom_text(data = subset(plot_data, is.na(MeanExpression)),
            aes(x = Tissue_grouped, y = Gene, label = "X"),
            color = "black", size = 3) + # Cross size and color
  scale_fill_gradientn(
    colors = c("#440154", "#31688E", "#35B779", "#FDE725"), # Gradient colors
    values = scales::rescale(c(0, 1, 2, 3)), # Define color breakpoints
    limits = c(0, 3), # Data limits
    oob = scales::squish, # Handle out-of-bound values
    na.value = "white", # White background for NA values
    labels = c("No", "Low", "Medium", "High") # Custom labels for the legend
  ) +
  labs(
    x = "",
    y = "",
    fill = "Mean\nexpression" # Update legend label for color
  ) +
  theme(
    panel.grid = element_blank(), 
    panel.background = element_rect(fill = "white"),
    panel.border = element_blank(), # Remove all borders
    axis.line.x = element_line(color = "black"), # Keep x-axis line
    axis.line.y = element_line(color = "black"), # Keep y-axis line
    axis.ticks.length = unit(0.1, "cm"), # Increase tick length
    axis.ticks.x = element_line(color = "black"), # Show x-axis ticks
    axis.ticks.y = element_line(color = "black"), # Show y-axis ticks
    axis.text.y = element_text(color = y_tick_colors),  # Assign colors to y-axis tick labels
    axis.text.x = element_text(color = "black", angle = 90, vjust = 0.5, hjust = 1, size = fontSize), # Larger x-axis labels
    axis.ticks.direction = "out", # Ticks face outward
    legend.position = "right"
  )
dev.off()


######################################### expression in Cells #########################################
candidateCARs_equivalent_Protein_Expression_res_list <- calcEquivalentExpression2(candidate_CARs, HPA_Protein_matrix_Cells_final_matrix) 

### Convert the list to a data frame
candidateCARs_equivalent_Protein_Expression_res_df_Cells <- as.data.frame(candidateCARs_equivalent_Protein_Expression_res_list)
colnames(candidateCARs_equivalent_Protein_Expression_res_df_Cells) <- names(candidateCARs_equivalent_Protein_Expression_res_list)
candidateCARs_equivalent_Protein_Expression_res_df_Cells = candidateCARs_equivalent_Protein_Expression_res_df_Cells[candidate_CARs]
colnames(candidateCARs_equivalent_Protein_Expression_res_df_Cells) <- candidate_CARs_display

candidateCARs_equivalent_Protein_Expression_res_df_Cells$Cell_grouped = rownames(candidateCARs_equivalent_Protein_Expression_res_df_Cells)
save(candidateCARs_equivalent_Protein_Expression_res_df_Cells, file = "/02.Input/HPA/version24_20241022/rna/candidateCARs_77_equivalent_Protein_Expression_res_df_Cells_v2.rdata")

Cells_reordered = c(
  "Heart muscle", "Neuronal", "Liver epithelial", "Alveolar", "Urinary epithelial", "Hematopoietic", "Endocrine", "Integumentary", "Endothelial/vascular", 
  "Gastrointestinal epithelial", "Immune", 
  "Female reproductive", "Male reproductive", 
  "Skeletal/smooth muscle", "General/glandular epithelial", "Glial", "Connective/stromal")

# Gather the data into long format
genes <- rev(candidate_CARs_display)
dotplot_data <- candidateCARs_equivalent_Protein_Expression_res_df_Cells %>%
  pivot_longer(
    cols = all_of(genes),
    names_to = "Gene",
    values_to = "Expression"
  )

################## plot cell Protein expression dotplot ################## 
# Calculate required metrics: zero ratio and mean expression for each Cell-Gene combination
plot_data <- dotplot_data %>%
  group_by(Cell_grouped, Gene) %>%
  summarise(
    ExprRatio = mean(Expression > 0, na.rm = TRUE), # Proportion of zeros
    MeanExpression = mean(Expression, na.rm = TRUE), # Average expression
    .groups = "drop"
  )

plot_data <- plot_data %>%
  mutate(
    Cell_grouped = factor(Cell_grouped, levels = Cells_reordered),
    Gene = factor(Gene, levels = genes) # Keep genes in column order
  )

# Create the plot
pdf_file <- paste0(fig_dir, "Figure3A_safety_landscape_Protein_Cell.pdf")
fig_width = 5.7
fig_height = 12.5
fontSize <- 10 # Larger font size for axis tick labels
pdf(pdf_file, onefile = FALSE, width = fig_width, height = fig_height)
ggplot(plot_data, aes(x = Cell_grouped, y = Gene)) +
  # Main heatmap with MeanExpression
  geom_tile(aes(fill = MeanExpression), width = 1, height = 1) +
  # Add a cross for NA values
  geom_text(data = subset(plot_data, is.na(MeanExpression)),
            aes(x = Cell_grouped, y = Gene, label = "X"),
            color = "black", size = 3) + # Cross size and color
  scale_fill_gradientn(
    colors = c("#440154", "#31688E", "#35B779", "#FDE725"), # Gradient colors
    values = scales::rescale(c(0, 1, 2, 3)), # Define color breakpoints
    limits = c(0, 3), # Data limits
    oob = scales::squish, # Handle out-of-bound values
    na.value = "white", # White background for NA values
    labels = c("No", "Low", "Medium", "High") # Custom labels for the legend
  ) +
  labs(
    x = "",
    y = "",
    fill = "Mean\nexpression" # Update legend label for color
  ) +
  theme(
    panel.grid = element_blank(), 
    panel.background = element_rect(fill = "white"),
    panel.border = element_blank(), # Remove all borders
    axis.line.x = element_line(color = "black"), # Keep x-axis line
    axis.line.y = element_line(color = "black"), # Keep y-axis line
    axis.ticks.length = unit(0.1, "cm"), # Increase tick length
    axis.ticks.x = element_line(color = "black"), # Show x-axis ticks
    axis.ticks.y = element_line(color = "black"), # Show y-axis ticks
    axis.text.y = element_text(color = y_tick_colors),  # Assign colors to y-axis tick labels
    axis.text.x = element_text(color = "black", angle = 90, vjust = 0.5, hjust = 1, size = fontSize), # Larger x-axis labels
    axis.ticks.direction = "out", # Ticks face outward
    legend.position = "right"
  )
dev.off()

```

# Figure S3. Grid plot showing mean safety score of candidate CARs (4 groups)
```{r}

total_candidate_CARs = length(candidate_CARs_display)

####### add mean efficacy 
candidate_CARs_max_expr_df = candidate_CARs_df[c("LogiCARs", "LogiCARs_display", "Efficacy_mean", "Group2")]
colnames(candidate_CARs_max_expr_df) = c("LogiCARs", "LogiCARs_display", "Efficacy", "Group")

####### calculate max RNA expression across tissues and cell types
# mean expression in each tissue
grouped_means <- candidateCARs_equivalent_RNA_Expression_res_df %>%
  group_by(Tissue_grouped) %>%
  summarise(across(1:total_candidate_CARs, \(x) mean(x, na.rm = TRUE)))
# max mean-expression across tissues
final_col_max <- apply(grouped_means[, -1], 2, max, na.rm = TRUE)
add_df <- as.data.frame(final_col_max)
colnames(add_df) = c("RNA (tissue)")
add_df$LogiCARs_display = rownames(add_df)
candidate_CARs_max_expr_df = merge(candidate_CARs_max_expr_df, add_df, by = "LogiCARs_display")

# mean expression in each cell type
grouped_means <- candidateCARs_equivalent_RNA_Expression_res_df %>%
  group_by(cell_group_custom) %>%
  summarise(across(1:total_candidate_CARs, \(x) mean(x, na.rm = TRUE)))
# max mean-expression across tissues
final_col_max <- apply(grouped_means[, -1], 2, max, na.rm = TRUE)
add_df <- as.data.frame(final_col_max)
colnames(add_df) = c("RNA (cell type)")
add_df$LogiCARs_display = rownames(add_df)
candidate_CARs_max_expr_df = merge(candidate_CARs_max_expr_df, add_df, by = "LogiCARs_display")


####### calculate max Protein expression across tissues and cell types
# max mean-expression across tissues
a <- candidateCARs_equivalent_Protein_Expression_res_df_Tissues
a$Tissue_grouped = NULL
final_col_max <- apply(a, 2, max, na.rm = TRUE)
final_col_max <- ifelse(is.infinite(final_col_max) & final_col_max < 0, NA, final_col_max)
add_df <- as.data.frame(final_col_max)
colnames(add_df) = c("Protein (tissue)")
add_df$LogiCARs_display = rownames(add_df)
candidate_CARs_max_expr_df = merge(candidate_CARs_max_expr_df, add_df, by = "LogiCARs_display")

# mean expression in each cell type
a <- candidateCARs_equivalent_Protein_Expression_res_df_Cells
a$Cell_grouped = NULL
final_col_max <- apply(a, 2, max, na.rm = TRUE)
final_col_max <- ifelse(is.infinite(final_col_max) & final_col_max < 0, NA, final_col_max)
add_df <- as.data.frame(final_col_max)
colnames(add_df) = c("Protein (cell type)")
add_df$LogiCARs_display = rownames(add_df)
candidate_CARs_max_expr_df = merge(candidate_CARs_max_expr_df, add_df, by = "LogiCARs_display")

candidate_CARs_max_expr_df[1:5,1:5]

candidate_CARs_max_expr_df <- candidate_CARs_max_expr_df[order(candidate_CARs_max_expr_df$Group, -candidate_CARs_max_expr_df$Efficacy), ]

rownames(candidate_CARs_max_expr_df) = candidate_CARs_max_expr_df$LogiCARs_display
candidate_CARs_max_expr_df0 <- candidate_CARs_max_expr_df
candidate_CARs_order = rownames(candidate_CARs_max_expr_df0)
candidate_CARs_max_expr_df = t(candidate_CARs_max_expr_df0[c("Efficacy", "RNA (tissue)", "RNA (cell type)", "Protein (tissue)", "Protein (cell type)")])
candidate_CARs_max_expr_df = as.data.frame(candidate_CARs_max_expr_df)

# Prepare data for the plot
candidate_CARs_max_expr_df$Metric <- rownames(candidate_CARs_max_expr_df) # Add Metric as a column
plot_data <- candidate_CARs_max_expr_df %>%
  pivot_longer(
    cols = -Metric, 
    names_to = "Gene", 
    values_to = "Value"
  ) %>%
  mutate(Value = as.numeric(Value)) # Ensure all values are numeric

# Separate data for dot plot (first row) and heatmap (rows 2-5)
dotplot_data <- plot_data %>% filter(Metric == "Efficacy")
dotplot_data$Gene = factor(dotplot_data$Gene, levels = candidate_CARs_order)
heatmap_data <- plot_data %>% filter(Metric != "Efficacy")
heatmap_data$Gene = factor(heatmap_data$Gene, levels = candidate_CARs_order)

# Create the plot
pdf_file <- paste0(fig_dir, "FigureS3_safety_landscape_RNA_and_Protein.pdf")
fig_width <- 16
fig_height <- 3.8
fontSize <- 10 # Larger font size for axis tick labels
pdf(pdf_file, onefile = FALSE, width = fig_width, height = fig_height)

ggplot() +
  # Dot plot for Efficacy
  geom_point(
    data = dotplot_data,
    aes(x = Gene, y = Metric, size = Value),
    color = "black"
  ) +
  scale_size_continuous(
    range = c(1, 5), # Dot size range
    limits = c(0, 1), # Scale legend from 0 to 1
    name = "Mean\nefficacy"
  ) +
  # Heatmap for rows 2-5
  geom_tile(
    data = heatmap_data,
    aes(x = Gene, y = Metric, fill = Value),
    width = 1,
    height = 1
  ) +
  # Add X for NA values
  geom_text(
    data = heatmap_data %>% filter(is.na(Value)),
    aes(x = Gene, y = Metric, label = "X"),
    color = "black",
    size = 3
  ) +
  scale_fill_gradientn(
    colors = c("#440154", "#31688E", "#35B779", "#FDE725"), # Gradient colors
    values = scales::rescale(c(0, 1, 2, 3)), # Define color breakpoints
    limits = c(0, 3), # Data limits
    oob = scales::squish, # Handle out-of-bound values
    na.value = "white", # White background for NA values
    name = "Max\nexpression",
    labels = c("No", "Low", "Medium", "High") # Custom labels for the legend
  ) +
  theme(
    panel.grid = element_blank(), 
    panel.background = element_rect(fill = "white"),
    panel.border = element_blank(), # Remove all borders
    axis.line.x = element_line(color = "black"), # Keep x-axis line
    axis.line.y = element_line(color = "black"), # Keep y-axis line
    axis.ticks.length = unit(0.1, "cm"), # Increase tick length
    axis.ticks.x = element_line(color = "black"), # Show x-axis ticks
    axis.ticks.y = element_line(color = "black"), # Show y-axis ticks
    axis.text.y = element_text(color = "black", size = fontSize), # Larger y-axis labels
    axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = fontSize, color = rev(y_tick_colors)), # Larger x-axis labels
    axis.ticks.direction = "out", # Ticks face outward
    legend.position = c(1.05, -0.6),
    legend.text = element_text(family = "Arial"), # Use a Unicode-compatible font: Arial sans
    plot.margin = margin(5, 100, 5, 5) # Add space on the right for the legend
    #legend.position = "right"
  ) +
  labs(title = "",
    x = "",
    y = "",
    size = "Efficacy", 
    fill = "Expression\nlevels")

dev.off()

```


