---
title: "Design paitent-specific logic-gated CARs for precision oncology"
author: "Tiangen Chang tiangen.chang@nih.gov"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# load required package
```{r, include=FALSE}

library(SeuratData)
library(SeuratDisk)
library(caret)
library(Seurat)
library(dplyr)
library(ggplot2)
library(dittoSeq)
library(SeuratWrappers)
library(patchwork)
library(RColorBrewer)
library(stringr)
library(tidyr)
library(data.table)
library(Matrix)
library(showtext) # this package enables R to show special characters like ≤
# Add a font and enable `showtext`
font_add(family = "Arial", regular = "path/to/arial.ttf")  # Replace with the path to your Arial font file
showtext_auto()

```


# set input and output directories & parameters
```{r}

data_dir <- paste0('/02.Input/')
processed_dir <- paste0('/03.Results/')
fig_dir = paste0('/03.Results/Figures/')
surface_genes = read.csv(file = "/02.Input/Constants/cell_surface_genes_wgene_and_clinical_uniqueStandardNames.txt", header = T) 
surface_genes = as.character(surface_genes[,1])

# Define custom colors for each group
group_colors <- c(
  "Patient-specific LogiCAR (N ≤ 3)" = "#08306b",  # Deep blue
  "General LogiCAR (N ≤ 3)" = "#2171b5",  # Dark blue
  "Kwon et al." = "#006d2c",      # Deep green
  "Dannenfelser et al." = "#41ab5d",  # Light green
  "Clinical (BRCA)" = "#cb181d",  # Deep red
  "Clinical (other tumors)" = "#fb6a4a"  # Light red (pink)
)

```

# load common functions
```{r}

source("00.Common_Functions.R")

```

# Read in and process top N-gene logic-gated CARs identified by LogiCAR Designer to standard format
```{r}

patients_prefix = c("GSE161529_patient0135_allTumorCells", "Bassez_CohortA_BIOKEY_10_allTumorCells", "GSE246613_Patient11_allTumorCells")
top_logiCARs_files_suffix = c("singlets", "doublets_merged", "triplets_merged")
keep_topN = 15

top_logiCARs_summary_list = list()
for (pp in patients_prefix){
  print(paste(pp, "in processing ..."))
  top_logiCARs_data_dir = paste0("/01.Scripts/Triplet_CART_designer/output/", pp, "/")
  
  ### extract top solutions with considering duplications (with N-1 genes the same)
  top_logiCARs_summary_df = data.frame()
  top_logiCARs_uniqueGenes = c()
  for (fn in top_logiCARs_files_suffix) {
    print(paste("  ", fn, "in processing ..."))
    top_logiCARs_df = read.csv(file = paste0(top_logiCARs_data_dir, pp, "_optimal_",fn,".txt"), sep = "\t", header = T)
    top_logiCARs_dt = as.data.table(top_logiCARs_df)
    # Remove brackets and split genes into a vector
    top_logiCARs_dt$Genes <- str_replace_all(top_logiCARs_dt$Genes, "\\[|\\]|\\'", "")
    top_logiCARs_dt[, GenesSplit := strsplit(Genes, ", ")]
    current_list = list()
    current_list[[1]] <- list(top_logiCARs_dt$GenesSplit[[1]], round(top_logiCARs_dt$Score[[1]],2), top_logiCARs_dt$LogicGates[[1]])
    current_list_len = 1
    for (row_i in 2:nrow(top_logiCARs_dt)){
      flag = 1
      Score_temp = round(top_logiCARs_dt$Score[[row_i]],2)
      if (fn != "singlets"){
        for (item_j in 1:current_list_len){
          diff_num = length(setdiff(current_list[[item_j]][[1]], top_logiCARs_dt$GenesSplit[[row_i]]))
          if ((diff_num < 2) & (Score_temp == current_list[[item_j]][[2]])){
            flag = 0
            break
          }
        }
      }
      if (flag == 1){
        current_list[[current_list_len+1]] <- list(top_logiCARs_dt$GenesSplit[[row_i]], Score_temp, top_logiCARs_dt$LogicGates[[row_i]])
        current_list_len = current_list_len + 1
        if (current_list_len == keep_topN){
          break
        }
      }
    }
    rows_used = row_i
    print(paste("  ", fn, rows_used))
    # Apply the conversion function to the data to get standard representation of logiCARs
    top_logiCARs_dt[, LogiCARs := mapply(convert_logic, LogicGates, Genes)]
    unique_genes <- unique(unlist(top_logiCARs_dt$GenesSplit[1:rows_used]))
    unique_genes <- gsub("_not", "", unique_genes)
    result_df_temp <- top_logiCARs_dt[1:rows_used, c("Genes", "LogiCARs", "Score", "GenesSplit")]
    geneNum = length(top_logiCARs_dt$GenesSplit[[1]])
    result_df_temp$geneNum = geneNum
    top_logiCARs_summary_df = rbind(top_logiCARs_summary_df, result_df_temp)
    top_logiCARs_uniqueGenes = unique(c(top_logiCARs_uniqueGenes, unique_genes))
  }
  ## convert - in gene names to _
  top_logiCARs_summary_df$LogiCARs = gsub("-", "_", top_logiCARs_summary_df$LogiCARs)
  top_logiCARs_summary_list[[pp]] = top_logiCARs_summary_df
}

```

# Load and process sampled normal cell RNA expression data (the LogiCAR Designer input)
```{r}

####### load HPA RNA Discovery data  31000 cells *  5448/2 genes
data_HPA_RNA = fread(file = paste0("/02.Input/HPA/version24_20241022/rna/HPA_LogiCAR_cell_surface_genes_wgene_and_clinical.txt"), sep = "\t", header = F) 
data_HPA_RNA_onlyData = data_HPA_RNA[3:nrow(data_HPA_RNA),,drop=F]
data_HPA_RNA_geneNAs = data_HPA_RNA_onlyData$V1
data_HPA_RNA_onlyData$V1 = NULL

# Step 1: Convert the data.table to a numeric matrix
data_HPA_RNA_onlyData <- as.matrix(data_HPA_RNA_onlyData)
rownames(data_HPA_RNA_onlyData) <- data_HPA_RNA_geneNAs
data_HPA_RNA_onlyData <- apply(data_HPA_RNA_onlyData, 2, as.numeric)  # Convert to numeric
# Step 2: Convert the numeric matrix to a sparse matrix
data_HPA_RNA_onlyData <- Matrix(data_HPA_RNA_onlyData, sparse = TRUE)
# Step 3: Transpose the sparse matrix
data_HPA_RNA_onlyData <- t(data_HPA_RNA_onlyData)
colnames(data_HPA_RNA_onlyData) <- data_HPA_RNA_geneNAs
data_HPA_RNA_onlyData_negated <- data_HPA_RNA_onlyData == 0
data_HPA_RNA_onlyData_negated <- as(data_HPA_RNA_onlyData_negated, "CsparseMatrix")
colnames(data_HPA_RNA_onlyData_negated) = paste0(data_HPA_RNA_geneNAs, "_not")
data_HPA_RNA_final_matrix <- cbind(data_HPA_RNA_onlyData, data_HPA_RNA_onlyData_negated)
## convert - in gene names to _
colnames(data_HPA_RNA_final_matrix) = gsub("-", "_", colnames(data_HPA_RNA_final_matrix))

```

# Load and process tumor cell RNA expression data (the LogiCAR Designer input)
```{r}

#### Bassez_CohortA_BIOKEY_10
dataset_name = "Bassez_2021_NatMed/CohortA/"
file_path <- paste0(processed_dir, dataset_name, "Bassez_2021_NatMed_CohortA_BIOKEY_10_LogiCAR_wgene_and_clinical.txt")
LogiCAR_input_df <- fread(file_path, sep = "\t", header = FALSE, data.table = FALSE, check.names = FALSE)
# Remove the first column and first three rows
numeric_data <- LogiCAR_input_df[-(1:3), -1]
# Convert the numeric matrix to a sparse matrix
numeric_matrix <- as.matrix(apply(numeric_data, 2, as.numeric))
Bassez_CohortA_BIOKEY_10_Smat <- Matrix(numeric_matrix, sparse = TRUE)
geneNAs_temp <- LogiCAR_input_df$V1[4:nrow(LogiCAR_input_df)]
sum(Bassez_CohortA_BIOKEY_10_Smat>0)/length(Bassez_CohortA_BIOKEY_10_Smat) # 0.0415483
Bassez_CohortA_BIOKEY_10_Smat <- t(Bassez_CohortA_BIOKEY_10_Smat)
colnames(Bassez_CohortA_BIOKEY_10_Smat) <- geneNAs_temp
Bassez_CohortA_BIOKEY_10_Smat_negated <- Bassez_CohortA_BIOKEY_10_Smat == 0
Bassez_CohortA_BIOKEY_10_Smat_negated <- as(Bassez_CohortA_BIOKEY_10_Smat_negated, "CsparseMatrix")
colnames(Bassez_CohortA_BIOKEY_10_Smat_negated) = paste0(geneNAs_temp, "_not")
Bassez_CohortA_BIOKEY_10_final_matrix <- cbind(Bassez_CohortA_BIOKEY_10_Smat, Bassez_CohortA_BIOKEY_10_Smat_negated)
## convert - in gene names to _
colnames(Bassez_CohortA_BIOKEY_10_final_matrix) = gsub("-", "_", colnames(Bassez_CohortA_BIOKEY_10_final_matrix))


#### GSE161529_patient0135
dataset_name = "GSE161529_patient0135"
file_path <- paste0(processed_dir, dataset_name, "_allTumorCells_LogiCAR_wgene_and_clinical.txt")
LogiCAR_input_df <- fread(file_path, sep = "\t", header = FALSE, data.table = FALSE, check.names = FALSE)
# Remove the first column and first three rows
numeric_data <- LogiCAR_input_df[-(1:3), -1]
# Convert the numeric matrix to a sparse matrix
numeric_matrix <- as.matrix(apply(numeric_data, 2, as.numeric))
GSE161529_patient0135_Smat <- Matrix(numeric_matrix, sparse = TRUE)
geneNAs_temp <- LogiCAR_input_df$V1[4:nrow(LogiCAR_input_df)]
sum(GSE161529_patient0135_Smat>0)/length(GSE161529_patient0135_Smat) # 0.0415483
GSE161529_patient0135_Smat <- t(GSE161529_patient0135_Smat)
colnames(GSE161529_patient0135_Smat) <- geneNAs_temp
GSE161529_patient0135_Smat_negated <- GSE161529_patient0135_Smat == 0
GSE161529_patient0135_Smat_negated <- as(GSE161529_patient0135_Smat_negated, "CsparseMatrix")
colnames(GSE161529_patient0135_Smat_negated) = paste0(geneNAs_temp, "_not")
GSE161529_patient0135_final_matrix <- cbind(GSE161529_patient0135_Smat, GSE161529_patient0135_Smat_negated)
## convert - in gene names to _
colnames(GSE161529_patient0135_final_matrix) = gsub("-", "_", colnames(GSE161529_patient0135_final_matrix))


#### GSE246613_Patient11
dataset_name = "GSE246613_Patient11"
file_path <- paste0(processed_dir, "GSE246613/", dataset_name, "_LogiCAR_wgene_and_clinical.txt")
LogiCAR_input_df <- fread(file_path, sep = "\t", header = FALSE, data.table = FALSE, check.names = FALSE)
# Remove the first column and first three rows
numeric_data <- LogiCAR_input_df[-(1:3), -1]
# Convert the numeric matrix to a sparse matrix
numeric_matrix <- as.matrix(apply(numeric_data, 2, as.numeric))
GSE246613_Patient11_Smat <- Matrix(numeric_matrix, sparse = TRUE)
geneNAs_temp <- LogiCAR_input_df$V1[4:nrow(LogiCAR_input_df)]
sum(GSE246613_Patient11_Smat>0)/length(GSE246613_Patient11_Smat) # 0.0415483
GSE246613_Patient11_Smat <- t(GSE246613_Patient11_Smat)
colnames(GSE246613_Patient11_Smat) <- geneNAs_temp
GSE246613_Patient11_Smat_negated <- GSE246613_Patient11_Smat == 0
GSE246613_Patient11_Smat_negated <- as(GSE246613_Patient11_Smat_negated, "CsparseMatrix")
colnames(GSE246613_Patient11_Smat_negated) = paste0(geneNAs_temp, "_not")
GSE246613_Patient11_final_matrix <- cbind(GSE246613_Patient11_Smat, GSE246613_Patient11_Smat_negated)
## convert - in gene names to _
colnames(GSE246613_Patient11_final_matrix) = gsub("-", "_", colnames(GSE246613_Patient11_final_matrix))

```


# Filtering top N-gene logic-gated CARs identified by LogiCAR Designer by safety
# Filter identified top logiCARs by safety (for same efficacy and N-1 same gene logiCARs, only keep the one with the highest safety)
```{r}

### function to filter LogiCARs to keep top N non-redundant targets
has_n_minus_1_common_genes <- function(genes1, genes2) {
  diff_genes <- setdiff(genes1, genes2)
  return((length(diff_genes) <2) & length(genes1) > 1)
}

### run the three patients one by one.
data_tumor_final_matrix_list = list(Bassez_CohortA_BIOKEY_10_allTumorCells = Bassez_CohortA_BIOKEY_10_final_matrix,
                                    GSE161529_patient0135_allTumorCells = GSE161529_patient0135_final_matrix,
                                    GSE246613_Patient11_allTumorCells = GSE246613_Patient11_final_matrix)

filtered_top_logiCARs_list = list()
for (pp in names(top_logiCARs_summary_list)){
  print(paste(pp, "in processing ..."))
  top_logiCARs_summary_df <- top_logiCARs_summary_list[[pp]]
  # efficacy RNA
  efficacy_res_list <- calcEquivalentExpression2(top_logiCARs_summary_df$LogiCARs, data_tumor_final_matrix_list[[pp]]) 
  # safety RNA
  safety_res_list <- calcEquivalentExpression2(top_logiCARs_summary_df$LogiCARs, data_HPA_RNA_final_matrix) 
  result_df = data.frame(LogiCARs = as.character(),
                         Efficay = as.numeric(),
                         Safety = as.numeric()
                         )
  # Loop through the names in safety_res_list
  for (le in names(safety_res_list)) {
    # Calculate Efficacy and Safety
    Efficacy <- mean(efficacy_res_list[[le]])
    Safety <- 1 - mean(safety_res_list[[le]])
    # Append the results as a new row to the data frame
    result_df <- rbind(result_df, data.frame(LogiCARs = le, Efficacy = Efficacy, Safety = Safety))
  }
  top_logiCARs_summary_df_scored = merge(top_logiCARs_summary_df, result_df, by = "LogiCARs")
  # Reduce the resolution of Score
  top_logiCARs_summary_df_scored[, Score := round(Score, 2)]
  ## Deduplicate rows by grouping rows with N-1 genes the same and the same Score
  filtered_rows <- list()
  for (gene_num in unique(top_logiCARs_summary_df_scored$geneNum)) {
    group <- top_logiCARs_summary_df_scored[geneNum == gene_num]
    keep_rows <- !logical(nrow(group)) # all TRUE
    for (i in 1:(nrow(group)-1)) {
      if (keep_rows[i]) {
        for (j in (i + 1):nrow(group)) {
          if (keep_rows[j] &&
              group$Score[i] == group$Score[j] &&
              has_n_minus_1_common_genes(group$GenesSplit[[i]], group$GenesSplit[[j]])) {
              if (group$Safety[i] >= group$Safety[j]) {
                keep_rows[j] <- FALSE
              } else {
                keep_rows[i] <- FALSE
                break
              }
            }
          }
        }
      }
    print(paste(gene_num, sum(keep_rows)))
    filtered_rows[[as.character(gene_num)]] <- group[keep_rows]
  }
  filtered_top_logiCARs <- rbindlist(filtered_rows)
  ## Keep only the top 10 rows per geneNum after deduplication
  filtered_top_logiCARs <- filtered_top_logiCARs[
    order(-Score, -Safety),
    head(.SD, 10),
    by = geneNum
  ]
  filtered_top_logiCARs = filtered_top_logiCARs[order(-filtered_top_logiCARs$Efficacy), ]
  filtered_top_logiCARs <- filtered_top_logiCARs[
    order(filtered_top_logiCARs$geneNum, -filtered_top_logiCARs$Efficacy),
  ]
  filtered_top_logiCARs_list[[pp]] <- filtered_top_logiCARs
}

save(filtered_top_logiCARs_list, file = "/01.Scripts/Triplet_CART_designer/output/Bassez_CohortA_BIOKEY_10_allTumorCells/filtered_top_logiCARs_v2.RData")

```

# Extract all candidate CAR targets and genes for comparison
```{r}

clinical_targets = c("MUC1", "CD274", "PROM1", "CD70", "ROR1", "CD80", "CD86", "TNFRSF10B", "FOLR1", "MPP2", "MUC16", "PMEL", "ROR2", "KDR", "EPHA2", "PSCA", "MET", "IL13RA2", "EPCAM", "EGFR", "FOLH1", "GPC3", "ERBB2", "MSLN", "CD276", "TM4SF1", "BSG", "CEACAM5", "ALPP", "GUCY2C", "TACSTD2")
### Top 10 previous bulk RNAseq derived computational combinations (exclude combinations with non-surface genes)
Computational_CAR_bulk <- c("FAP & EPHA2_not", "FAP & MET_not", "OR2B6 & EPHA2_not", "OR2B6 & MET_not", "SLC1A4 & SSTR1_not", "ADAM12 & KDR_not", "EPHB3 & FZD5_not", "ADAM12 & GPC3_not", "MUC1 & LYVE1_not", "SLC4A5 & PTH1R_not")
### Top 5 previous scRNAseq derived computational combinations for AND, OR, NOT gates respectively
Computational_CAR_sc = c("SLC39A6 & SPINT2", "TACSTD2 & TSPAN13", "CD24 & TM4SF1", "CD9 & SLC39A6", "BCAM & SPINT2",
                        "BAMBI | PRLR", "NECTIN4 | PRLR", "BAMBI | NECTIN4", "LY6K | PRLR", "PRLR | VTCN1",
                        "CLDN3 & TSPAN8_not", "SERINC2 & TSPAN8_not", "CLDN3 & PIGR_not", "PLPP2 & MUC4_not", "SERINC2 & PIGR_not"
                        )
other_targets <- c()

### only keep top 5 for each N
for (pp in names(filtered_top_logiCARs_list)){
  filtered_top_logiCARs <- filtered_top_logiCARs_list[[pp]]
  filtered_top_logiCARs <- filtered_top_logiCARs[
    order(-Score, -Safety),
    head(.SD, 5),
    by = geneNum
  ]
  filtered_top_logiCARs_list[[pp]] <- filtered_top_logiCARs
}

filtered_targets_df_list <- list()
for (pp in names(filtered_top_logiCARs_list)){
  filtered_top_logiCARs <- filtered_top_logiCARs_list[[pp]]
  logiCAR_combos <- filtered_top_logiCARs$LogiCARs
  targets_vec <- c(logiCAR_combos, clinical_targets, Computational_CAR_bulk, Computational_CAR_sc, other_targets)
  logiCAR_combos_df = as.data.frame(filtered_top_logiCARs)[c("LogiCARs", "geneNum")]
  logiCAR_combos_df$Group = "LogiCAR Designer"
  Clinical_CARs_df = data.frame(LogiCARs = clinical_targets)
  Clinical_CARs_df$geneNum = 1
  Clinical_CARs_df$Group = "Clinical Targets"
  Computational_CARs_df <- data.frame(LogiCARs = c(Computational_CAR_bulk, Computational_CAR_sc))
  Computational_CARs_df$geneNum = 2
  Computational_CARs_df$Group = c(rep("Previous Computational (bulk)", length(Computational_CAR_bulk)), rep("Previous Computational (sc)", length(Computational_CAR_sc)))
  Other_CARs_df <- if (length(other_targets) > 0) {
    data.frame(
      LogiCARs = other_targets,
      geneNum = NA,
      Group = "Other Targets"
    )
  } else {
    data.frame(
      LogiCARs = character(0),
      geneNum = numeric(0),
      Group = character(0)
    )
  }
  targets_df <- rbind(logiCAR_combos_df, Clinical_CARs_df, Computational_CARs_df, Other_CARs_df)
  
  ## get unique single genes from above vector
  cleaned_combos <- gsub("[|&()]", "", targets_vec)  # Remove |, &, (, )
  cleaned_combos <- gsub("_not", "", cleaned_combos)    # Remove _not
  all_genes <- unlist(strsplit(cleaned_combos, "\\s+"))
  unique_genes <- unique(all_genes)
  unique_CARargets_measured <- intersect(unique_genes, unique_genes) # allGenes_tumor allGenes_measured
  ## get targets that all genes present in HPA protein measurement
  # Function to check if all genes in a logic combination are in unique_CARargets_measured
  is_valid_combo <- function(combo, valid_genes) {
    # Extract genes from the logic combination
    genes <- unlist(strsplit(combo, "[|&() ]+"))  # Split by logical operators and spaces
    genes <- gsub("_not", "", genes)    # Remove _not
    genes <- genes[genes != ""]  # Remove empty strings
    # Check if all genes are in valid_genes
    all(genes %in% valid_genes)
  }
  filtered_indices = sapply(targets_vec, is_valid_combo, valid_genes = unique_CARargets_measured)
  filtered_targets_vec <- targets_vec[filtered_indices]
  filtered_targets_df <- targets_df[filtered_indices, ]
  
  filtered_targets_df_list[[pp]] <- filtered_targets_df
}

```


# Figure 5ABC. Barplot rank BRCA CAR targets based on  efficacy on each patient
```{r}

### run the three patients one by one.
data_tumor_final_matrix_list = list(Bassez_CohortA_BIOKEY_10_allTumorCells = Bassez_CohortA_BIOKEY_10_final_matrix,
                                    GSE161529_patient0135_allTumorCells = GSE161529_patient0135_final_matrix,
                                    GSE246613_Patient11_allTumorCells = GSE246613_Patient11_final_matrix)

patient_in_test = "GSE161529_patient0135" # "Bassez_CohortA_BIOKEY_10", "GSE246613_Patient11", "GSE161529_patient0135"
name_for_data = "GSE161529_patient0135_allTumorCells" #  "Bassez_CohortA_BIOKEY_10_allTumorCells", "GSE246613_Patient11_allTumorCells" , "GSE161529_patient0135_allTumorCells"

filtered_targets_df <- filtered_targets_df_list[[name_for_data]]
data_tumor_final_matrix <- data_tumor_final_matrix_list[[name_for_data]]
### add "general LogiCARs" and remove low-safety "Previous Computational" and "Clinical Targets"
existing_logiCARs_with_safety = read.csv("/01.Scripts/Triplet_CART_designer/output/safety90/filtered_top_logiCARs_efficacy_safety_17Datasets.csv")
general_LogiCARs_df <- existing_logiCARs_with_safety[(existing_logiCARs_with_safety$Group == "LogiCAR Designer") & (existing_logiCARs_with_safety$geneNum <= 3), c("LogiCARs", "geneNum", "Group")]
general_LogiCARs_df$Group <- "general_LogiCARs"
filtered_targets_df <- rbind(filtered_targets_df, general_LogiCARs_df)
existing_CARs_indices <- existing_logiCARs_with_safety$Group %in% c("Previous Comptutional (bulk)","Clinical Targets","Previous Comptutional (sc)")
safe_indices <- existing_logiCARs_with_safety$Safety >= 0.9
safe_existing_CARs = existing_logiCARs_with_safety$LogiCARs[existing_CARs_indices & safe_indices]
remove_indices <- (!(filtered_targets_df$LogiCARs %in% safe_existing_CARs)) &(filtered_targets_df$Group %in% c("Previous Comptutional (bulk)","Clinical Targets","Previous Comptutional (sc)"))
filtered_targets_df <- filtered_targets_df[!remove_indices, ]

## identify duplicated CAR targets in different groups
duplicated_CARs = names(table(filtered_targets_df$LogiCARs)[table(filtered_targets_df$LogiCARs) > 1])
filtered_targets_df[filtered_targets_df$LogiCARs %in% duplicated_CARs, ]
## mark duplicates
filtered_targets_df$LogiCARs_display <- filtered_targets_df$LogiCARs
for (cc in duplicated_CARs) {
  # Find the indices of the current duplicated LogiCARs value
  indices <- which(filtered_targets_df$LogiCARs == cc)
  # Add "*", "**", etc., based on the occurrence count
  for (i in seq_along(indices)) {
    if (i > 1) {  # Skip the first occurrence
      filtered_targets_df$LogiCARs_display[indices[i]] <- paste0(cc, strrep("*", i - 1))
    }
  }
}

###For GSE161529; Patient: 0135
#filtered_targets_df[filtered_targets_df$LogiCARs == "(CCR8 | TACSTD2) & EMP2", ] <- c("(ITM2B_not | TACSTD2) & EMP2", 3, "LogiCAR Designer", "(ITM2B_not | TACSTD2) & EMP2")
###For Bassez CohortA: delete the row that filtered_targets_df$LogiCARs=="ALPP"
#filtered_targets_df <- filtered_targets_df[filtered_targets_df$LogiCARs != "ALPP", ] # not measured in Bassez CohortA

# efficacy RNA
efficacy_res_list <- calcEquivalentExpression2(filtered_targets_df$LogiCARs, data_tumor_final_matrix) 
length(efficacy_res_list)
length(efficacy_res_list[[1]])
names(efficacy_res_list) <- filtered_targets_df$LogiCARs_display

result_df = data.frame(LogiCARs_display = as.character(),
  Efficay = as.numeric()
  )
# Loop through the names in safety_res_list
for (le in names(efficacy_res_list)) {
  Efficacy <- mean(efficacy_res_list[[le]])
  result_df <- rbind(result_df, data.frame(LogiCARs_display = le, Efficacy = Efficacy))
}
result_df$LogiCARs_display = NULL

filtered_targets_result_df = cbind(filtered_targets_df, result_df)
filtered_targets_result_df1 = filtered_targets_result_df[filtered_targets_result_df$Group != "general_LogiCARs",]
filtered_targets_result_df2 = filtered_targets_result_df[filtered_targets_result_df$Group == "general_LogiCARs",]
## Keep only the top 5 rows per geneNum based on Efficacy for general_LogiCARs
filtered_targets_result_df2 <- filtered_targets_result_df2 %>%
  filter(Group == "general_LogiCARs") %>%  # Filter rows where Group is "general_LogiCARs"
  group_by(geneNum) %>%                   # Group by geneNum
  arrange(desc(Efficacy)) %>%             # Sort by Efficacy in descending order
  slice_head(n = 5) %>%                   # Select the top 5 rows for each group
  ungroup()                               # Ungroup the data frame (optional)
filtered_targets_result_df = rbind(filtered_targets_result_df1, filtered_targets_result_df2)

# To replace _not at the end of words with ! as a prefix 
filtered_targets_result_df$LogiCARs_display <- gsub(
  pattern = "\\b(\\w+)_not\\b", 
  replacement = "!\\1", 
  filtered_targets_result_df$LogiCARs_display
)
## convert _ in gene names to -
filtered_targets_result_df$LogiCARs_display = gsub("_", "-", filtered_targets_result_df$LogiCARs_display)
filtered_targets_result_df$geneNum = NULL
filtered_targets_result_df$LogiCARs <- filtered_targets_result_df$LogiCARs_display

filtered_targets_result_df$Group[filtered_targets_result_df$Group == "Clinical Targets"] = "Clinical (other tumors)"
filtered_targets_result_df$Group[filtered_targets_result_df$LogiCARs %in% c("MUC1", "CD274", "PROM1", "CD70", "ROR1", "PSCA", "MET", "EPCAM", "EGFR", "FOLH1", "ERBB2", "MSLN", "CD276", "TACSTD2")] = "Clinical (BRCA)"
filtered_targets_result_df$Group[filtered_targets_result_df$Group == "LogiCAR Designer"] = "Patient-specific LogiCAR (N \u2264 3)"
filtered_targets_result_df$Group[filtered_targets_result_df$Group == "Previous Comptutional (bulk)"] = "Dannenfelser et al."
filtered_targets_result_df$Group[filtered_targets_result_df$Group == "Previous Comptutional (sc)"] = "Kwon et al."
filtered_targets_result_df$Group[filtered_targets_result_df$Group == "general_LogiCARs"] = "General LogiCAR (N \u2264 3)"
unique(filtered_targets_result_df$Group)

##### delete general LogiCARs
filtered_targets_result_df <- filtered_targets_result_df[filtered_targets_result_df$Group != "General LogiCAR (N \u2264 3)", ]

######################## plot barplot of mean efficacy of targets across 17 datasets with errorbars and scatters, 
# Order by desired Group order and descending Mean_Efficacy within each group
group_order <- c("Patient-specific LogiCAR (N \u2264 3)", "General LogiCAR (N \u2264 3)", "Kwon et al.", "Dannenfelser et al.", "Clinical (BRCA)", "Clinical (other tumors)")
filtered_targets_result_df <- filtered_targets_result_df %>%
  arrange(factor(Group, levels = group_order), desc(Efficacy))

filtered_targets_result_df$Group = factor(filtered_targets_result_df$Group, levels = group_order)
filtered_targets_result_df$LogiCARs = factor(filtered_targets_result_df$LogiCARs, levels = filtered_targets_result_df$LogiCARs)

# Barplot with scatter points colored by Dataset
pdf_file <- paste0(fig_dir,paste0("Figure5_efficacy_",patient_in_test,"_ranking_barplot.pdf"))
fig_width = 9 # 11
fig_height = 4.5
fontSize = 1.2
pdf(pdf_file, onefile=FALSE, width = fig_width,height=fig_height)
ggplot(filtered_targets_result_df, aes(x = LogiCARs, y = Efficacy * 100, fill = Group)) +
  geom_bar(stat = "identity", position = position_dodge(width = 0.8)) +
  labs(
    title = "",
    x = "",
    y = "Efficacy (%)",
    fill = "Group"   # Remove legend title for fill
  ) +
  scale_fill_manual(
    values = group_colors
  ) +
  scale_y_continuous(limits = c(0, 100)) +
  theme(
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "white"),
    axis.line.y = element_line(color="black"),
    axis.line.x = element_line(color="black"),
    axis.ticks.y = element_line(color="black"),
    axis.ticks.x = element_line(color="black"),
    axis.text.y = element_text(color="black"),
    panel.border = element_blank(),
    axis.text.x = element_text(color = "black", angle = 90, vjust = 0.5, hjust = 1),
    legend.position = "none", # c(1.1, 0.6) "none" c(0.8, 0.6)
    legend.text = element_text(family = "Arial"), # Use a Unicode-compatible font: Arial sans
    plot.margin = margin(5, 5, 5, 5) # margin(5, 200, 5, 5)
  )
dev.off()

```



# Figure S5. UMAP visualization of coverage of tumor cells by different CARs (in Dataset: GSE161529; Patient: 0135)
```{r}

### GSE161529 (patient: 0135)
dataset_name = 'GSE161529'
file_prefix = "seu_Tissue"
data_dir <- paste0("/02.Input/", dataset_name, "/")
processed_dir <- data_dir

seurat_obj <- readRDS(file = paste0(processed_dir,"Bassez_CohortA_BIOKEY_10_tumorCells.rds"))

################## Renormalize 
expression_matrix <- GetAssayData(seurat_obj, layer = "counts")
surface_expression_matrix <- expression_matrix[rownames(expression_matrix) %in% surface_genes, ]
seurat_obj_subset <- CreateSeuratObject(counts = surface_expression_matrix, meta.data = seurat_obj@meta.data)

##### renormalization and UMAP based ONLY on relevant CAR genes (and clinical targets)
gene_names <- c("FOLR1", "PROM1", "CD276", "NECTIN4", "LDLRAD3", "ITM2B", "TACSTD2", "EMP2")
#gene_names <- unique(c(gene_names, clinical_targets))
seurat_obj_subset <- NormalizeData(seurat_obj_subset)
seurat_obj_subset <- ScaleData(seurat_obj_subset, features = gene_names)
seurat_obj_subset <- RunPCA(seurat_obj_subset, features = gene_names)
seurat_obj_subset <- RunUMAP(seurat_obj_subset, dims = 1:3)  # Use the first 10 PCs

expression_data <- GetAssayData(seurat_obj_subset, assay = "RNA", layer = "data")
seurat_obj_subset$CAR_FOLR1 <- (expression_data["FOLR1", ] > 0)
seurat_obj_subset$CAR_PROM1 <- (expression_data["PROM1", ] > 0)
seurat_obj_subset$CAR_CD276 <- (expression_data["CD276", ] > 0)
seurat_obj_subset$CAR_NECTIN4 <- (expression_data["NECTIN4", ] > 0)
seurat_obj_subset$CAR_LDLRAD3 <- (expression_data["LDLRAD3", ] > 0)

seurat_obj_subset$CAR_ITM2B_not_TACSTD2_EMP2 <- ((expression_data["ITM2B", ] == 0 | expression_data["TACSTD2", ] > 0) & expression_data["EMP2", ] > 0)

CAR_targets = c("CAR_FOLR1", "CAR_PROM1", "CAR_CD276", "CAR_NECTIN4", "CAR_LDLRAD3", "CAR_ITM2B_not_TACSTD2_EMP2") 
my.colors <- c("Expressed" = "red", "Not expressed" = "grey")
for (ct in CAR_targets){
  seurat_obj_subset$CAR_target = ifelse(seurat_obj_subset[[ct]] == TRUE, "Expressed", "Not expressed")
  DimPlot(
    seurat_obj_subset,
    group.by = "CAR_target",  # Do not group by any metadata
    label = FALSE,   # Remove labels for groups
    repel = TRUE,    # Adjust labels to avoid overlap if present
    raster = FALSE   # Higher quality plot without rasterization
  ) +
    scale_color_manual(values = my.colors) + 
    ggtitle("") +                         
    labs(x = "UMAP 1", y = "UMAP 2") +
    scale_x_continuous(breaks = scales::pretty_breaks(n = 4), labels = scales::number_format(accuracy = 1)) + 
    scale_y_continuous(breaks = scales::pretty_breaks(n = 4), labels = scales::number_format(accuracy = 1)) +
    NoLegend()                      
  ggsave(
    filename = paste0(fig_dir, "FigureS5_UMAP_Bassez_CohortA_BIOKEY_10_tumor_",ct,".pdf"), 
    height = 90, 
    width = 90, 
    units = "mm", 
    device = "pdf" 
  )
}

```
